DNA replication in primary hepatocytes without the six-subunit ORC

Version 1

Abstract

Abstract

The six subunit ORC is essential for initiation of DNA replication in eukaryotes. Cancer cell-lines in culture can survive and replicate DNA replication after genetic inactivation of individual ORC subunits, ORC1, ORC2 or ORC5. In primary cells, ORC1 was dispensable in the mouse liver for endo-reduplication, but this could be explained by the ORC1 homolog, CDC6, substituting for ORC1 to restore functional ORC. Here, we have created mice with a conditional deletion of ORC2, which does not have a homolog. Although mouse embryo fibroblasts require ORC2 for proliferation, mouse hepatocytes synthesize DNA in cell culture and endo-reduplicate in vivo without ORC2. Mouse livers endo-reduplicate after simultaneous deletion of ORC1 and ORC2 both during normal development and after partial hepatectomy. Since endo-reduplication initiates DNA synthesis like normal S phase replication these results unequivocally indicate that primary cells, like cancer cell lines, can load MCM2-7 and initiate replication without ORC.

Introduction

DNA replication during normal mitotic cycles or during endo-reduplication is initiated by the six subunit ORC, five of which have AAA+ATPase domains and form a ring shaped complex that binds DNA and bends it (Bell and Stillman, 1992, Neuwald et al., 1999, Dhar et al., 2001a, Clarey et al., 2006, Bleichert et al., 2015). In cooperation with another ATPase, CDC6, and with CDT1, ORC helps load MCM2-7 double hexamers at or near origins of replication(Costa and Diffley, 2022, Stillman, 2022, Hu and Stillman, 2023). To initiate replication (or endo-reduplication), the MCM2-7 hexamer associates with additional proteins to form the functional CMG helicase that is essential for unwinding the double stranded chromosomal DNA, so that the resulting single-stranded DNAs can serve as templates for copying by DNA polymerases. In the lower eukaryotes, various ORC subunits have been consistently found to be essential for DNA replication and cell proliferation (Bell et al., 1993, Foss et al., 1993, Micklem et al., 1993, Loo et al., 1995, Semple et al., 2006). In mammalian cells, knockdown of individual ORC subunits have also been reported to stop cell proliferation (Chou et al., 2021, Prasanth et al., 2004). In humans, mutations in Orc1, Orc4, Orc6, Cdt1, and Cdc6 cause microcephalic primordial dwarfism resembling Meier-Gorlin syndrome (Bicknell et al., 2011b, Bicknell et al., 2011a, Guernsey et al., 2011).

There have, however, been isolated reports that suggest that under unusual circumstances, DNA replication can initiate in eukaryotes in the absence of the full ORC. First hypomorphic mutation made in the Orc2 gene in HCT116 colon cancer cells decreased ORC2 protein levels by 90% and cells were still viable and could proliferate (Dhar et al., 2001b). Deletion of the Orc1 gene in Drosophila permitted several rounds of replication in the resulting larvae and pupae (Park and Asano, 2008). A few human cancer cell lines have been created by CRISPR-Cas9 mediated genome engineering where ORC1, ORC2 and ORC5 cannot be detected by regular immunoblots and yet the cell lines survive, proliferate and replicate DNA with the normal complement of origins of replication (Shibata et al., 2016, Shibata and Dutta, 2020). However, the Orc2 gene is essential for viability and in the Orc2Δ cell line a very minimal level (<0.1% of wild type levels) of a truncated protein can be detected that reacts with anti-ORC2 antibody and co-immunoprecipitates with ORC3 (Chou et al., 2021). Given that WT cells have about 150,000 molecules of ORC2, even if this truncated protein is functional ORC2, ∼150 molecules of the protein would be expected to load MCM2-7 double hexamers on at least 50,000 origins of replication. This possibility would be consistent with current models of MCM2-7 loading only if ORC was highly catalytic and one ORC hexamer was capable of loading ∼667 MCM2-7 hexamers (Costa and Diffley, 2022, Stillman, 2022, Hu and Stillman, 2023).

To address whether the entire ORC holocomplex is essential for DNA replication initiation in primary mammalian cells, we have reported a conditional mutation in the ORC1 gene of mouse (Okano-Uchida et al., 2018). Conditional deletion of Orc1 (using tissue-specific Cre drivers) revealed that the gene was essential for normal mitotic cell division in intestinal epithelial cells, but that endoreduplication in tissues that are known to become polyploid, like placental trophoblasts and the liver hepatocytes, was unaffected. Because the ORC1 protein is a functional ATPase with sequence homology with CDC6 (Saha et al., 1998, Williams et al., 1997), one possibility that remained was that the mouse CDC6 protein substituted for ORC1 in the ORC holocomplex, producing enough functional ORC. Therefore, we decided to make a conditional mutation in another subunit of ORC, Orc2, that does not have the Walker A or B motifs that would make it a functional ATPase and that is not homologous to Cdc6. We designed the mutation such that even if an N terminally truncated protein is expressed from a downstream methionine from an alternately spliced mRNA the AAA+ like domain will suffer a significant deletion. This will allow us to test whether the mouse liver survival and endoreduplication after loss of ORC1 is seen even after loss of ORC2. It will also test whether combining the conditional mutations of ORC1 and ORC2 will allow the liver cells to survive and endoreduplicate. If hepatocyte survival and endoreduplication is seen in these mice, then it would be hard to explain this by the substitution of two ORC subunits by the CDC6 protein, and we would gain support for the hypothesis that in rare circumstances even untransformed mammalian cells are capable of loading enough MCM2-7 in the absence of ORC to permit extensive DNA replication.

Results

ORC2 is essential for embryonic development and proliferation of mouse embryo fibroblasts

Mice with Lox P sites inserted flanking exons 6 and 7 of mouse Orc2 were purchased from Cyagen Biosciences Inc. (Fig. 1A). Genotyping with primers F2 and R2 distinguished the loxP marked allele from the WT allele (Fig. 1B). In a cross of Orc2f/+ mice, there was no significant decrease in the yield of Orc2f/+ or Orc2f/f compared to Orc2+/+ mice (Fig. 1C), suggesting that the insertion of the loxP sites in the introns of Orc2 did not impair the function of ORC2. Recombination between the lox2 sites would delete the coding exons 6 and 7 of Orc2, which removes amino acids L111-V150 (Fig. 1D). If a transcript is expressed that skips exons 6 and 7, then the resulting protein mutates A110 to V110 and then throws the protein coding sequence out of frame (Fig 1D, ΔORC2), so that the bulk of the 576 amino acid ORC2 protein, including the AAA+ like and WH domains (K230-A576) are not expressed. The AAA+ like domain and the WH domain are key elements for ORC2 assembly into ORC and for ORC function. Even in the remote possibility that a truncated protein is expressed due to alternative splicing and translation initiation from an internal methionine, the next methionine is at M301, so that half of the protein including 70 amino acids of the 239 amino acid AAA+-like domain (K230-G469) would be deleted.

Generation of Orc2f/fmice and ORC2 KO MEFs.

A. Scheme of introduced loxP sites in Orc2 locus.

B. Representative picture of genotyping of offspring coming from Orc2f/+ crossed with Orc2f/+.

C. The ratio of observed to expected animals coming from Orc2f/+ crossed with Orc2f/+.

D. Schematic of the ORC2 protein and the DeltaORC2 protein produced after deletion of exons 6 and 7. A110 is mutated to V110 and then the protein goes out of frame.

E. Validation of Orc2 deletion 3 days after Adeno cre transduction.

F. Western blot of ORC2 protein 5 days after Adeno cre transduction. 10 or indicated ul of lysate loaded/lane as written on the top

G. MTT assay of WT and Orc2f/f MEFs without and with Adeno cre transduction.

H. Western blot of ORC2 protein 5 and 15 days after Adeno cre transduction.

The Sox2-Cre allele expresses active Cre during embryonic development. Crossing the Orc2f/f mice with Sox2-Cre mice resulted in no Orc2Δ/Δ embryos at E7.5 days onwards suggesting that embryonic deletion of Orc2 is lethal (Table 1). The near expected number of Orc2+/Δ embryos suggests that hemizygosity of Orc2 can still support viability.

Mouse embryo fibroblasts (MEF) were obtained from Orc2f/f E13.5 day embryos and cultured in vitro. 3 days after infection with an Adeno virus expressing Cre, effective recombination between the loxP sites is seen, resulting in a genotype showing that exons 6 and 7 have been deleted in most of the MEF (Orc2Δ) (Fig. 1E). Consistent with this, at 5 days after adeno-Cre infection, the ORC2 protein is not detected in the MEF population (Fig. 1F). The dilution of the undeleted MEF lysate on the same blot suggests, that if any ORC2 protein is expressed in the MEF after adeno-Cre infection, it is <10% of that in the undeleted MEF. The proliferation rate of the MEF were measured by MTT assays. Even in the Orc2+/+ MEF, the infection with adeno-Cre decreased proliferation a little (the orange line compared to the blue line in Fig. 1G). However, for Orc2f/f MEF infection with adeno-Cre impairs proliferation even further (yellow line compared to black line in Fig. 1G). Furthermore, when the MEF were cultured for 15 days, the surviving cells express significant levels of ORC2 (Fig. 1H), suggesting that the MEFs that had not undergone the Cre-mediated deletion take over the culture.

Taken together, these results suggest that Cre mediated deletion of exons 6 and 7 of Orc2 leads to early embryonic lethality and impairs the proliferation of normal diploid MEFs in culture.

Knockout of Orc2 in developing mouse liver makes ORC2 protein undetectable in hepatocytes and yet supports most of normal development and endoreduplication

Mice carrying one copy of the Albumin promoter driven Cre gene (Alb-Cre) express the Cre recombinase specifically in hepatocytes (Postic et al., 1999). When the Orc2f/f mice are crossed with mice carrying Alb-Cre (shortened below as Alb), the Cre recombinase is expected to promote the recombination mediated deletion of exons 6 and 7 of Orc2 in the hepatocytes. The Alb+/--Orc2f/f mice were viable and fertile (Fig. 2A, 2B). Crossing such mice yielded the expected numbers of Alb+/--Orc2f/f mice, though there was a drop in the yield of Alb+/+-Orc2f/f mice. One allele of Alb-Cre is sufficient to express enough Cre to carry out homozygous deletion of the Orc2 allele in the hepatocytes, so that the results suggest that deletion of Orc2 in the hepatocytes did not impair viability. The partial lethality of the Alb+/+ (HOM) mice could be due to the toxicity of high dose of Cre recombinase expressed, something that has been noted by other groups (Schmidt et al., 2000, Loonstra et al., 2001, Schmidt-Supprian and Rajewsky, 2007, Naiche and Papaioannou, 2007, Janbandhu et al., 2014). Because the Alb-Cre is expressed only in the hepatocytes and not in other cells in the liver, we isolated hepatocytes from these livers by growing them in culture (schematically shown in Fig. 4A) to determine whether the ORC2 protein was decreased (Fig. 2C). The ORC2 protein was significantly decreased in all five of the Alb+/-- Orc2f/f mice, showing consistent and near 100% effect of the deletion in the hepatocytes. Western blots for several other ORC subunits, and CDC6 protein showed that the loss of ORC in the same hepatocytes did not significantly decrease these proteins, though MCM2 and MCM3 (two subunits of the MCM2-7 helicase component) were decreased by 50% (Fig. 2D). The body weights of the Alb, Orc2f/f mice were smaller than in the Orc2f/f animals of both sexes, though the liver size and liver size normalized to body weight was significantly smaller only in the females (Fig. 2E-G). There was some elevation of the circulating liver enzymes in the mice where Alb-Cre is expressed to delete the Orc2 gene, suggesting that there is some impairment of liver function (Supp. Fig. S1).

Development of liver in Orc2 KO mice.

A. Scheme of Alb+/--Orc2f/f ROSA26stop-EYFP crossed with Alb+/--Orc2f/f ROSA26stop-EYFP (All mice are with ROSA26stop-EYFP and so we do not include this in the genotypes below)

B. The ratio of observed to expected animals coming from A.

C. Western blot of hepatocytes from Orc2 f/f and Orc2 f/f Alb-Cre +/- animals. Tubulin was used as loading control.

D. Quantification of the Western blots of hepatocyte lysates from Orc2 f/f (without Alb-cre) mice and the same genotype but with Alb-Cre to show the levels of other key replication initiation proteins in the ORC2 KO hepatocytes.

E. Average body weight of Orc2f/f and Alb-Orc2f/f animals.

F. Average liver weight of Orc2f/f and Alb-Orc2f/f animals.

G. Average liver to body weight ratio of Orc2f/f and Alb-Orc2f/f animals.

H. Representative H&E staining of liver tissue from Orc2f/f (WT) and Alb-Orc2f/f (KO) animals. Both panels at same scale.

I. Quantification of hepatocyte nuclear size in Orc2f/f and Alb-Orc2f/f animals.

J. Quantification of hepatocyte nuclear size in Orc2f/f and Alb-Orc2f/f female mice.

K. Quantification of hepatocytes nuclear size in Orc2f/f and Alb-Orc2f/f male mice.

*p < 0.05, **p < 0.01, two-tailed Student’s t test.

The livers of mice can grow by proliferation (increasing the number of cells) or by hypertrophy of cells (larger cells with endoreduplicated nuclei). Histological examination of the livers revealed that the Orc2 deletion was accompanied by the presence of fewer (∼50% of WT levels), but larger nuclei and cells (Fig. 2H and 2I). The Orc2 deleted hepatocytes had significantly larger nuclei in both males and females (Fig. 2J, K). The larger nuclei in hepatocytes with Orc2 deletion also seen in other contexts, e.g. in the pre-resection hepatocytes with Orc2 deletion (Fig. 5E, 0 hr point) and in the Cre expressing, Orc1f/+, Orc2f/f that have an intact Orc1 gene but delete both alleles of Orc2 (Supp. Fig. S3B, green bars compared to blue bars).

To determine whether the larger nuclei were generated by endoreduplication, we isolated nuclei from hepatocytes and stained them with the DNA staining dye, DRAQ5, followed by flow cytometry (Fig. 3A). Both male and female Alb-Cre, Orc2f/f hepatocytes showed a decrease in 4n DNA content and an increase in 16n DNA content (Fig. 3B-D). The expression of Cre would also delete a lox-stop-lox element upstream from EYFP and lead to the expression of EYFP. Indeed, when we gated on EYFP positive cells, the 8n and 16N DNA content increase was much more evident relative to the EYFP negative cells, while the 2N DNA containing nuclei were decreased (Fig. 3E-G). Thus Orc2 deletion promotes endo-reduplication, a result similar to what was noted when Orc1 was deleted in mouse livers (Okano-Uchida et al., 2018).

Endoreduplication in the Orc2 liver conditional knock-out animals.

A. Experimental design.

B-D. Quantification of nuclei ploidy in the livers of Orc2f/f ROSA26stop-EYFP and Alb-Orc2f/f ROSA26stop-EYFP animals.

E-G. Quantification of nuclei ploidy for EYFP low (negative) and high (positive) primary hepatocytes.

*p < 0.05, **p < 0.01, ***p < 0.001, two-tailed Student’s t test.

DNA replication of Alb-Cre+/--Orc2f/f hepatocytes in vitro

Primary hepatocytes isolated from 8-10 weeks old mouse liver can replicate their DNA and proliferate for a limited time in vitro. We isolated such hepatocytes from Alb-cre-/- (WT) and Alb-cre+/-livers, both from Orc2f/f mice (Fig. 4A). Genotyping reveals that over 90% of the cells from the Alb-cre+/-livers have successfully deleted Orc2 exons 6 and 7 (Fig. 4B), while immunoblotting shows that ORC2 protein expression is also significantly decreased in these cells (Fig. 3C). EdU labeling in vitro for 2 hrs, showed that the EYFP positive cells (where Cre recombinase has been active) can incorporate EdU (Fig. 4C), although the total number of nuclei that incorporate EdU is decreased to about 30% of that seen in the ORC2 wild type hepatocytes (Fig. 4D). Thus hepatocytes in culture can continue to replicate DNA in the absence of detectable ORC2 protein.

Orc2 KO primary hepatocytes are viable and can incorporate EdU in vitro.

A. Experimental design

B. Genotyping and western blotting of hepatocytes

C. Representative picture of EdU, EYFP and DAPI staining on the Orc2 WT (Orc2f/f) and KO (Orc2f/f Alb-Cre) primary hepatocytes.

D. The percentage of EdU positive nuclei from Orc2 WT or Orc2 KO primary hepatocytes.

*p < 0.05, two-tailed Student’s t test.

DNA replication in Alb-Cre+/--Orc2f/f hepatocytes during liver regeneration in vivo

Liver DNA synthesis peaks around 36-48 hours, and the liver regenerates to nearly 50% of its original weight within 2-3 days following partial hepatectomy. Such regeneration involves both normal mitotic DNA replication/cell division and endoreduplication accompanied by hypertrophy of the cells. We therefore tested whether deletion of the Orc2 gene in the hepatocytes adversely affects liver regeneration after partial hepatectomy in 8-14 weeks old mice (Fig. 5A). Nearly two-thirds of the liver is removed surgically and the mice allowed to recover for 36-48 hrs before harvesting the regenerated liver.

The ORC2 mutant livers regenerate after partial hepatectomy.

A. Schematic of the experiment.

B. Body weight of the Orc2f/f ROSA26stop-EYFP mice without (-/-) or with Alb-Cre (-/+) before partial hepatectomy.

C. Liver weight of the mice in B after liver regeneration.

D. Regenerated Liver to pre-hepatectomy body weight ratio of the mice in B.

E. H&E stain of Orc2f/f ROSA26stop-EYFP livers with intact Orc2 (Alb-cre -/-, N=3) or Orc2 knockout (Alb-cre+/-, N=7). Scale bar: 25 μm.

F. Quantitation of nuclear counts per field (76,000 um2). Six images were taken for each liver. 0 hr (pre-resection). 36 hr (post-regeneration).

G. EdU incorporation of indicated livers. EYFP marks cells where Cre has been expressed. Scale bar: 25 μm.

H. Percent EdU+ nuclei counted in 1882 and 825 nuclei in the Cre- and Cre+ livers, respectively.

I. Nuclear size of indicated livers. 0 hr (pre-resection). 36 hr (post-regeneration). Mean and S.D from about 40-70 nuclei, *p < 0.05, ****p < 0.0001, unpaired two-tailed Student’s t test is used

Although the liver weight after regeneration was smaller than that of WT livers, the liver:body weight ratio was similar in the livers with Orc2 deletion and Orc2 WT (Fig. 5B-D). The H&E stain of the regenerated liver shows that the liver cells are larger and have larger nuclei in the liver with Orc2 deletion (Fig. 5E). In these experiments, both before and after regeneration, the livers with the Orc2 deletion have ∼50% nuclei compared to the wild type livers (Fig. 5F). Finally, labeling the livers by injection of EdU in the mice 3-4 hrs before harvesting, shows that the EYFP positive cells (indicating activity of Cre recombinase) were proficient in synthesizing DNA and incorporating EdU (Fig. 5G, H). Nearly 100% of the hepatocytes were positive for EYFP, suggesting that the Cre recombinase was active in the vast majority of the hepatocytes (Fig. 5G), and consistent with the complete depletion of Orc2 protein from the hepatocytes of all mice with that genotype (Fig. 3C). Thus the 30% of the liver cells that were incorporating EdU in the Cre expressing livers (Fig. 5H) were doing so in the absence of Orc2 protein. Also, we noted several pairs of EdU positive nuclei (marked by arrows in Fig. 5G) whose relative positions suggest that they are sisters born from the same mitosis, suggesting that some EdU positive cells can go through mitosis in the EYFP positive cells. Finally, the nuclei were significantly larger in the Cre active Orc2f/f hepatocytes than in the hepatocytes without Cre, both before and after regeneration (Fig. 5I).

We did not explore why the EYFP protein is mostly nuclear in hepatocytes in culture (Fig. 4C) and mostly cytoplasmic in hepatocytes in the liver tissue (Fig. 5G, 7G), but speculate that differences in signaling pathways or fixation techniques between the two conditions contribute to this difference.

Viable mice with endoreduplicated hepatocyte nuclei in Alb-Orc1f/f Orc2f/f mice

We have reported that conditional deletion of Orc1 in developing mouse livers still allowed livers to develop and induced premature endoreduplication, suggesting that significant DNA synthesis can occur in liver cells that are genetically deleted of Orc1. We bred the Orc1f/f mice with Orc2f/f mice to obtain mice where both Orc1 and Orc2 are floxed but Alb-cre is not present (Fig. 6A, B). Even though Cre was not expressed in these livers, for unknown reasons there was a decrease in the percentage of progeny when Orc2 was floxed in the liver (with or without Orc1 being floxed), but not when just Orc1 was floxed (Fig. 6C). These mice were then bred with Orc2f/f Alb-cre+/+ mice, and the resulting Orc1f/+, Orc2f/f, Alb-cre+/- mice intercrossed to get Orc1f/f, Orc2f/f, Alb-Cre+/- mice where the Orc1 and Orc2 alleles are conditionally deleted in hepatocytes, and the deletion is accompanied by activation of EYFP expression. Immunoblotting of isolated hepatocytes showed that in four out of four mice expressing the Cre recombinase the ORC1 and ORC2 proteins were decreased significantly (Fig. 6D). Immunoblotting of some of the other replication proteins showed no decrease of other ORC subunits, CDC6 and two of the MCM2-7 subunits (Fig. 6E). The ORC3 protein, which did not change when Orc2 was deleted (Fig. 2D) was surprisingly elevated in Fig. 6E when Orc1 and Orc2 were both deleted. We do not know why this is the case, and it was not seen consistently in all four animals. Similarly, the 50% decrease of MCM2 and MCM3 that was seen in the absence of ORC2 (Fig. 2D), disappeared when both Orc1 and Orc2 are deleted.

Endoreduplication in liver of Orc1 Orc2 DKO animals.

A-B. Breeding schemes to obtain conditional double flox animals.

C. The ratio of observed to expected animals coming from B. Orc1 = all animals with Orc1f/f ROSA26stop-EYFP, Orc2 = all animals with Orc2f/f ROSA26stop-EYFP, Orc1 Orc2 = all animals with Orc1f/f Orc2f/f ROSA26stop-EYFP genotype. This was before introduction of Alb-Cre.

D. Immunoblot of hepatocytes from WT (Orc1f/f Orc2f/f) and DKO (Orc1f/f Orc2f/f Alb-cre+/-) mice to show that ORC1 and ORC2 are depleted in the DKO cells.

E. Quantitation of immunoblots to show that levels of other key initiation protein subunits are not decreased in the DKO mice hepatocytes.

F. Average body, liver weight and their ratio for WT and DKO animals.

G. Representative H&E staining of liver tissue from male WT and DKO animals.

H. Quantification of hepatocyte nuclear size in the WT and DKO animals.

I. Quantification of nuclei ploidy for EYFP low (negative) and high (positive) primary hepatocytes from DKO mice.

The female Alb-Orc1f/f Orc2f/f mice were significantly smaller in size, with smaller livers and decreased liver to body weight ratio (Fig. 6F, Supp. Fig. 2A), but the male mice were relatively less affected (Fig. 6F). We have also observed ∼50% lethality of double knock-out female mice within the first month of life, but the remaining 50% survive beyond 4 months of age (Supp. Fig. 2C). H&E staining showed that the double knockout (DKO) male livers have fewer but larger cells with significantly larger nuclei (Fig. 6G, 6H). Interestingly, this phenotype is even more marked in the female liver (Supp. Fig. 2B). By flow cytometry the EYFP positive cells as evidence of Cre recombinase activity had distinctly more polyploidization than the EYFP negative cells (Fig. 6I). Taken together the results reveal that hepatocytes can synthesize DNA by endoreduplication to produce very large nuclei and very large cells so that the liver size is not hugely decreased, even in the absence of two subunits of ORC. Female mice that have a deletion in two subunits of ORC suffer more morbidity and mortality than male mice.

DNA replication in Alb-Cre+/--Orc1f/f Orc2f/f hepatocytes during liver regeneration in vivo

Finally, partial hepatectomy was performed in the livers of male mice with DKO of Orc1 and Orc2. As can be seen even before partial hepatectomy the body weights were not significantly different from WT mice (Fig. 7A). Post regeneration, the liver weights and liver/body weight ratios were not decreased in the DKO mice compared to the WT mice (Fig. 7B, C). H&E sections showed that the hepatocyte nuclei were larger and hepatocyte nuclear density lower in the DKO livers both pre-hepatectomy and in the regenerated livers 36 hr post-hepatectomy (Fig. 7D-F). EdU labeling for 3-4 hrs before harvesting the livers showed that despite the DKO, the EYFP positive cells showed extensive DNA synthesis (Fig. 7G). Even though nearly 100% of the hepatocytes are EYFP positive (Fig. 7G) and ORC1 and ORC2 proteins depleted in the hepatocytes (Fig. 6D), 15 % of nuclei stained with EdU (Fig. 7H), a percentage that was higher in the DKO regenerating livers than in the WT livers, suggesting that the endoreduplication that accompanies liver regeneration can occur even after deletion of two of the six subunits of ORC.

Endoreduplication in the liver of the DKO mice during liver regeneration.

A-C. Body weight pre-resection, liver weight post-regeneration and regenerated liver to body weight ratio in mice with indicated genotypes. Black bars: 4 WT males (Orc1f/f Orc2f/f ROSA26stop-EYFP mice without Alb-Cre). White bars: 6 DKO males (Orc1f/f Orc2f/f ROSA26stop-EYFP mice with Alb-Cre +/-). No significant difference between the two groups using two tailed Student t-test.

D: H&E stain of WT (N=4) or DKO mice (N=6). Scale bar: 50 μm.

E. Quantitation of hepatocyte nuclear size post regeneration. WT: black bars. DKO: white bars. 0 hr (pre-resection). 36 hr (post-regeneration). Five-six images were taken for each liver. About 120-200 nuclei are counted.

F: Quantitation of hepatocyte nuclear density post regeneration. WT: black bars. DKO: white bars. 0 hr (pre-resection). 36 hr (post-regeneration). Five-six images were taken for each liver.

G. Micrographs of EdU, DAPI and EYFP imaging of livers with indicated genotypes post regeneration. Scale bar: 20 μm. WT in the top row, DKO in the bottom row.

H. Quantitation of EdU positive nuclei post regeneration. WT: black bar. DKO: white bar. Five-six images were taken for each liver. *p < 0.05, ****p < 0.0001, unpaired two-tailed Student’s t test were used

Discussion

The six subunit ORC is important for recruiting MCM2-7, the core of the replicative DNA helicase and is essential for all forms of DNA replication in eukaryotic cells(Costa and Diffley, 2022, Stillman, 2022, Hu and Stillman, 2023). Yet, it has been possible to select cancer cell-lines that have mutations in Orc1, Orc2 or Orc5 and do not express detectable levels of the proteins and yet load MCM2-7 to the chromatin and replicate their DNA as they proliferate in culture (Shibata et al., 2016, Shibata and Dutta, 2020). This paradox suggested that there may be an alternative way to load sufficient MCM2-7 on chromatin and support DNA replication, at least in cancer cells in culture. In addition, there have been two instances where Orc1 has been mutationally removed in Drosophila (Park and Asano, 2008) and in mouse hepatocytes or placental trophoblasts (Okano-Uchida et al., 2018), where extensive DNA replication (particularly endo-reduplication in hepatocytes and trophoblasts) has persisted, suggesting that an alternative way of loading MCM2-7 may be available in certain unique cell-cycles that are different from the normal DNA replication that occurs during normal mitotic growth of diploid cells.

Five of the six subunits of ORC (ORC1-5) associate to form a ring-shaped structure with interactions between the WH domains and the AAA+ like domains of the five subunits (Bleichert et al., 2015). In mammalian cells only ORC1 and ORC4 have the intact Walker A and B motifs that are necessary for the molecules to act as AAA+ ATPases (Giordano-Coltart et al., 2005). Though this has never been demonstrated experimentally, it is theoretically possible that in the absence of ORC1 in flies or in mouse livers, another related AAA+ ATPase, intimately involved in MCM2-7 loading, CDC6, can substitute for ORC1 to reconstitute a functional 5 subunit ring-shaped ORC-like structure that executes its function (Takeda et al., 2005, Bell, 2017). We therefore wished to test whether removal of a second subunit of ORC, ORC2, that does not have obvious ATPase activity and not much homology to CDC6, would still permit endoreduplication in mouse livers. In the cancer cells, we had removed the initiator methionine of the Orc2 gene (Shibata et al., 2016). Although 99.9% of the ORC2 protein disappeared, an N terminally truncated form of ORC2 protein was expressed in the mutant cells at 0.1% of the wild type level (Chou et al., 2021). Thus, in this case we took the additional precaution to design a mutation such that even if a truncated protein was expressed from an internal methionine downstream from the mutation site, such a protein will be missing a significant part of its AAA+ATPase domain. Anyhow, in Western blots we do not see any evidence of such a truncated protein being expressed in the genetically altered MEFs or hepatocytes. In addition, we wondered whether the simultaneous genetic removal of two subunits of ORC, Orc1 and Orc2, would successfully obliterate all forms of DNA replication, including endoreduplication.

Our results show that while Orc2 is genetically essential for viability of early embryos in utero and mouse embryo fibroblasts in vitro, mutational inactivation of Orc2, or combined inactivation of Orc1 and Orc2, does not significantly inhibit development of the mouse liver. Adult, viable mice are produced. Albumin expression is activated very early in hepatoblasts in the 7-8 somite stage and the albumin mRNA can be seen in the hepatic primordium emerging from the gut at E9.5 days (Gualdi et al., 1996). 3H-thymidine incorporation can be measured in the liver as late as days 7-14 postnatally (Tilghman and Belayew, 1982), so that many cycles of DNA replication and cell proliferation are expected to occur in hepatoblasts and hepatocytes after the activation of Alb-Cre and subsequent homozygous knockout of Orc2 (or Orc1) in the hepatocytes.

The presence of functional livers in the mutant (but viable) animals, suggested that the homozygous knockout of these genes did not have a profound effect on liver development, as would have been expected if the cells were as sensitive as MEFs in culture to the loss of the ORC holocomplex. We have been unable to find antibodies that will recognize mouse Orc1 or Orc2 proteins in immunohistochemistry on tissues and so decided to calculate how many cell divisions have to occur after the Orc1 or Orc2 genes are deleted in the embryonic mouse hepatocytes (see Methods and Table 2). The calculations suggest that the Orc2 deleted livers and male DKO livers underwent at least 18 cell divisions, while the female DKO livers underwent at least 15 cell divisions since E9.5 (Table 2, bottom row).

If we allow a generous 3 cell divisions for the Cre recombinase to definitively excise the Floxed genes (consistent with what we see in MEFs in culture), this means that 15 cell divisions in the Orc2 deleted livers and male DKO livers occurred after the gene(s) was/were deleted. Once a gene is deleted, each cell division decreases the corresponding protein at least by half, so that 6 cell divisions would dilute the targeted protein to <1.5% of the wild type levels. Thus even after the ORC2 (or ORC! And ORC2 in the DKO mice) decreased to <1.5% the WT level, the Orc2 knockout livers (male or female) and the male DKO livers could execute at least 9 more rounds of replication and cell division. Note the amount of the relevant ORC subunit would continue to decrease by half with each further division. Also, that the levels of ORC are likely to decrease faster if the protein is actively degraded or if more cell divisions are necessary to counter any development-related apoptosis.

This suggests that mitotic DNA replication of diploid cells is not completely eliminated at least in the Orc2-/- or male DKO livers after loss of the targeted ORC proteins. Consistent with this, we see EdU incorporation in EYFP+ hepatocytes, in vitro, which, however, could be from a terminal endo-reduplication event. More tantalizing is the presence of paired EYFP positive EdU-positive nuclei during the regeneration after partial hepatectomy in 8-14 week old mice (Fig. 5G). Such paired nuclei are usually seen with daughter nuclei that have just been born from mitosis, suggesting that there were at least a few cells that underwent complete DNA replication and mitosis even though the hepatocytes had undergone 15 rounds of cell division after deletion of the targeted ORC gene(s).

The female DKO (but not the Orc2-/-) livers seem more sensitive to the depletion of both ORC1 and ORC2 since ∼12 cell divisions in the female DKO livers occur after allowing 3 cell division for the targeted genes to be definitively deleted. Thus, after the cells have reached <1.5 % of WT levels of ORC2 or ORC1 protein, they can still execute at least 6 more cell divisions. Consistent with this sex specific difference, the male mice appear to be more tolerant of deletion of Orc2 gene or both Orc1 and Orc2 genes, with the livers reaching larger sizes than in the female mice. Plasma testosterone levels reach 50% of adult levels in the first day after birth before declining (to rise again at puberty at 4-5 weeks age), suggesting that enough androgens are present during the proliferative phase of liver development (Bell, 2018). Sex-specific changes in gene expression have been noted to begin in the liver by 3 weeks postnatally (Conforto and Waxman, 2012). Thus, it is possible that androgens directly or indirectly stimulate the division or hypertrophy of hepatocytes to ameliorate the negative effects on liver mass due to the loss of ORC.

In contrast to diploid DNA replication in hepatocytes, we have shown that the livers with a genetic deletion of Orc2 (or Orc1 and Orc2) undergo endo-reduplication more easily during normal development. Such DNA synthesis is also clearly seen during liver regeneration after partial hepatectomy in the Orc2-/- or Orc1-/-, Orc2-/- livers. Endoreduplication differs from normal replication in that although the cycles of DNA replication (S phase) are separated by a G1 like phase (G phase) they are not separated by an intervening mitosis and cytokinesis, and is commonly seen during normal development in plants and animals (Shu et al., 2018). However, all evidence suggests that the biochemical mechanism of initiating DNA replication is the same in normal mitotic replication and endoreduplication. For example, cyclin E/cdk kinase activity and MCM2-7 protein association with the chromatin oscillate between G phase and S phase) during Drosophila endoreduplication (Lilly and Duronio, 2005) (Su and O’Farrell, 1998). Similarly, in the mouse, cyclin E and CDC6 promote, while geminin inhibits endoreduplication (Welch, 1992) (Bermejo et al., 2002, Gonzalez et al., 2006) (de Renty et al., 2014). Thus, the persistence, and in fact acceleration, of endoreduplication in the mouse livers in the absence of ORC2 and/or ORC1 suggests that there must be an alternate way to recruit sufficient MCM2-7 to the chromatin to support at least the three cycles of endoreduplication required to produce a 16 N nucleus from a 2 N embryonic nucleus during normal development, and to support about two cycles of endoreduplication during liver regeneration.

The acceleration and persistence of endoreduplication even when both Orc1 and Orc2 are genetically inactivated, makes it unlikely that a make-shift ORC-like complex is being assembled for loading the MCM2-7 proteins when two out of the five subunits in the ring are missing. Our results are virtually identical to what was observed when we conditionally deleted Orc1 (Okano-Uchida et al., 2018). There too, endoreduplication in the liver cells not only persisted but was induced prematurely during development. One possible explanation of the greater endoreduplication in both our papers is that mitosis may be arrested earlier in development by G2 DNA damage checkpoints activated by incomplete licensing and replication of the genome in the absence of ORC. As a result, endoreduplication cycles could begin earlier in development resulting in greater endoreduplication.

Is it possible that incomplete deletion of Orc2 or Orc1 genes in the hepatocytes allows enough liver cells to still carry an Orc2 or Orc1 gene to support DNA replication? This is very unlikely, because the in vitro DNA replication experiment with hepatocytes shows that although the ORC2 protein is virtually undetectable in immunoblots and there is evidence of Cre activity in 100% of the hepatocytes, the number of nuclei incorporating EdU in culture are decreased to only 30% the wild type level. Furthermore, in the partial hepatectomy experiments, nearly 100% of the cells are positive for EYFP, no ORC2 (or ORC1 protein in the DKO) is detectable in immunoblots of isolated hepatocytes, suggesting high penetrance in the expression of the Cre recombinase in the hepatocytes. Yet the liver succeeds in endoreduplicating to reach nearly normal liver size and EdU incorporation is seen in 35% (Fig. 5H) and 15% (Fig. 7G) of hepatocytes from the Orc2 KO and DKO mice, respectively.

Although the liver size in the Alb-Cre, Orc2f/f mice is not significantly decreased relative to body weight, the liver function tests suggest some impairment of liver function. We cannot yet attribute this deficit of liver function to the decrease in number of cells, or to excess endoreduplication, because ORC subunits are also known to be important for epigenetic control of gene expression (Vermeulen et al., 2010). Future experiments will determine whether the liver pathology could be secondary to epigenetic dysregulation of genes important for liver function. The double knockout female mice are sicker, and here some of the explanation may lie in the smaller liver, but here again epigenetic dysregulation in the absence of ORC cannot be ruled out as a potential cause for the morbidity.

Interestingly, the deletion of Orc2 or of Orc1+Orc2, in the hepatocytes does not consistently increase or decrease any of the other ORC subunits or the examined proteins downstream of ORC like CDC6 or two of the subunits of MCM2-7. Thus, a change in any of these proteins is unlikely to be the explanation for how the hepatocytes license enough origins to support replication during development and regeneration in the absence of the ORC holocomplex. The hepatocyte studies say that the cancer cells are not unique in their ability to bypass the requirement of two ORC subunits. Even primary cells can sometimes replicate DNA in the absence of the ORC holocomplex.

eLife Assessment

This descriptive manuscript builds on prior research showing that the elimination of Origin Recognition Complex (ORC) subunits does not halt DNA replication. The authors use various methods to genetically remove one or two ORC subunits from specific tissues and observe continued replication, though it may be incomplete. The replication appears to be primarily endoreduplication, indicating that ORC-independent replication may promote genome reduplication without mitosis. Despite similar findings in previous studies, the paper provides convincing genetic evidence in mice that liver cells can replicate and undergo endoreduplication even with severely depleted ORC levels. While the mechanism behind this ORC-independent replication remains unclear, the study lays the groundwork for future research to explore how cells compensate for the absence of ORC and to develop functional approaches to investigate this process. The reviewers agree that this valuable paper would be strengthened significantly if the authors could delve a bit deeper into the nature of replication initiation, potentially using an origin mapping experiment. Such an exciting contribution would help explain the nature of the proposed new type of Mcm loading, thereby increasing the impact of this study for the field at large.

Reviewer #1 (Public review):

The origin recognition complex (ORC) is an essential loading factor for the replicative Mcm2-7 helicase complex. Despite ORC's critical role in DNA replication, there have been instances where the loss of specific ORC subunits has still seemingly supported DNA replication in cancer cells, endocycling hepatocytes, and Drosophila polyploid cells. Critically, all tested ORC subunits are essential for development and proliferation in normal cells. This presents a challenge, as conditional knockouts need to be generated, and a skeptic can always claim that there were limiting but sufficient ORC levels for helicase loading and replication in polyploid or transformed cells. That being said, the authors have consistently pushed the system to demonstrate replication in the absence or extreme depletion of ORC subunits.

Here, the authors generate conditional ORC2 mutants to counter a potential argument with prior conditional ORC1 mutants that Cdc6 may substitute for ORC1 function based on homology. They also generate a double ORC1 and ORC2 mutant, which is still capable of DNA replication in polyploid hepatocytes. While this manuscript provides significantly more support for the ability of select cells to replicate in the absence or near absence of select ORC subunits, it does not shed light on a potential mechanism.

The strengths of this manuscript are the mouse genetics and the generation of conditional alleles of ORC2 and the rigorous assessment of phenotypes resulting from limiting amounts of specific ORC subunits. It also builds on prior work with ORC1 to rule out Cdc6 complementing the loss of ORC1.

The weakness is that it is a very hard task to resolve the fundamental question of how much ORC is enough for replication in cancer cells or hepatocytes. Clearly, there is a marked reduction in specific ORC subunits that is sufficient to impact replication during development and in fibroblasts, but the devil's advocate can always claim minimal levels of ORC remaining in these specialized cells.

The significance of the work is that the authors keep improving their conditional alleles (and combining them), thus making it harder and harder (but not impossible) to invoke limiting but sufficient levels of ORC. This work lays the foundation for future functional screens to identify other factors that may modulate the response to the loss of ORC subunits.

This work will be of interest to the DNA replication, polyploidy, and genome stability communities.

Reviewer #2 (Public review):

This manuscript proposes that primary hepatocytes can replicate their DNA without the six-subunit ORC. This follows previous studies that examined mice that did not express ORC1 in the liver. In this study, the authors suppressed expression of ORC2 or ORC1 plus ORC2 in the liver.

Comments:

(1) I find the conclusion of the authors somewhat hard to accept. Biochemically, ORC without the ORC1 or ORC2 subunits cannot load the MCM helicase on DNA. The question arises whether the deletion in the ORC1 and ORC2 genes by Cre is not very tight, allowing some cells to replicate their DNA and allow the liver to develop, or whether the replication of DNA proceeds via non-canonical mechanisms, such as break-induced replication. The increase in the number of polyploid cells in the mice expressing Cre supports the first mechanism, because it is consistent with few cells retaining the capacity to replicate their DNA, at least for some time during development.

(2) Fig 1H shows that 5 days post infection, there is no visible expression of ORC2 in MEFs with the ORC2 flox allele. However, at 15 days post infection, some ORC2 is visible. The authors suggest that a small number of cells that retained expression of ORC2 were selected over the cells not expressing ORC2. Could a similar scenario also happen in vivo?

(3) Figs 2E-G shows decreased body weight, decreased liver weight and decreased liver to body weight in mice with recombination of the ORC2 flox allele. This means that DNA replication is compromised in the ALB-ORC2f/f mice.

(4) Figs 2I-K do not report the number of hepatocytes, but the percent of hepatocytes with different nuclear sizes. I suspect that the number of hepatocytes is lower in the ALB-ORC2f/f mice than in the ORC2f/f mice. Can the authors report the actual numbers?

(5) Figs 3B-G do not report the number of nuclei, but percentages, which are plotted separately for the ORC2-f/f and ALB-ORC2-f/f mice. Can the authors report the actual numbers?

(6) Fig 5 shows the response of ORC2f/f and ALB-ORC2f/f mice after partial hepatectomy. The percent of EdU+ nuclei in the ORC2-f/f (aka ALB-CRE-/-) mice in Fig 5H seems low. Based on other publications in the field it should be about 20-30%. Why is it so low here? The very low nuclear density in the ALB-ORC2-f/f mice (Fig 5F) and the large nuclei (Fig 5I) could indicate that cells fire too few origins, proceed through S phase very slowly and fail to divide.

(7) Fig 6F shows that ALB-ORC1f/f-ORC2f/f mice have very severe phenotypes in terms of body weight and liver weight (about on third of wild-type!!). Fig 6H and 6I, the actual numbers should be presented, not percentages. The fact that there are EYFP negative cells, implies that CRE was not expressed in all hepatocytes.

(8) Comparing the EdU+ cells in Fig 7G versus 5G shows very different number of EdU+ cells in the control animals. This means that one of these images is not representative. The higher fraction of EdU+ cells in the double-knockout could mean that the hepatocytes in the double-knockout take longer to complete DNA replication than the control hepatocytes. The control hepatocytes may have already completed DNA replication, which can explain why the fraction of EdU+ cells is so low in the controls. The authors may need to study mice at earlier time points after partial hepatectomy, i.e. sacrifice the mice at 30-32 hours, instead of 40-52 hours.

(9) Regarding the calculation of the number of cell divisions during development: the authors assume that all the hepatocytes in the adult liver are derived from hepatoblasts that express Alb. Is it possible to exclude the possibility that pre-hepatoblast cells that do not express Alb give rise to hepatocytes? For example the cells that give rise to hepatoblasts may proliferate more times than normal giving rise to a higher number of hepatoblasts than in wild-type mice.

(10) My interpretation of the data is that not all hepatocytes have the ORC1 and ORC2 genes deleted (eg EYFP-negative cells) and that these cells allow some proliferation in the livers of these mice.

Reviewer #3 (Public review):

Summary:

The authors address the role of ORC in DNA replication and that this protein complex is not essential for DNA replication in hepatocytes. They provide evidence that ORC subunit levels are substantially reduced in cells that have been induced to delete multiple exons of the corresponding ORC gene(s) in hepatocytes. They evaluate replication both in purified isolated hepatocytes and in mice after hepatectomy. In both cases, there is clear evidence that DNA replication does not decrease at a level that corresponds with the decrease in detectable ORC subunit and that endoreduplication is the primary type of replication observed. It remains possible that small amounts of residual ORC are responsible for the replication observed, although the authors provide arguments against this possibility. The mechanisms responsible for DNA replication in the absence of ORC are not examined.

Strengths:

The authors clearly show that there are dramatic reductions in the amount of the targeted ORC subunits in the cells that have been targeted for deletion. They also provide clear evidence that there is replication in a subset of these cells and that it is likely due to endoreduplication. Although there is no replication in MEFs derived from cells with the deletion, there is clearly DNA replication occurring in hepatocytes (both isolated in culture and in the context of the liver). Interestingly, the cells undergoing replication exhibit enlarged cell sizes and elevated ploidy indicating endoreduplication of the genome. These findings raise the interesting possibility that endoreduplication does not require ORC while normal replication does.

Weaknesses:

There are two significant weaknesses in this manuscript. The first is that although there is clearly robust reduction of the targeted ORC subunit, the authors cannot confirm that it is deleted in all cells. For example, the analysis in Fig. 4B would suggest that a substantial number of cells have not lost the targeted region of ORC2. Although the western blots show stronger effects, this type of analysis is notorious for non-linear response curves and no standards are provided. The second weakness is that there is no evaluation of the molecular nature of the replication observed. Are there changes in the amount of location of Mcm2-7 loading that is usually mediated by ORC? Does an associated change in Mcm2-7 loading lead to the endoreduplication observed? After numerous papers from this lab and others claiming that ORC is not required for eukaryotic DNA replication in a subset of cells, we still have no information about an alternative pathway that could explain this observation.

The authors frequently use the presence of a Cre-dependent eYFP expression as evidence that the ORC1 or ORC2 genes have been deleted. Although likely the best visual marker for this, it is not demonstrated that the presence of eYFP ensures that ORC2 has been targeted by Cre. For example, based on the data in Fig. 4B, there seems to be a substantial percentage of ORC2 genes that have not been targeted while the authors report that 100% of the cells express eYFP.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

This descriptive manuscript builds on prior research showing that the elimination of Origin Recognition Complex (ORC) subunits does not halt DNA replication. The authors use various methods to genetically remove one or two ORC subunits from specific tissues and observe continued replication, though it may be incomplete. The replication appears to be primarily endoreduplication, indicating that ORC-independent replication may promote genome reduplication without mitosis. Despite similar findings in previous studies, the paper provides convincing genetic evidence in mice that liver cells can replicate and undergo endoreduplication even with severely depleted ORC levels. While the mechanism behind this ORC-independent replication remains unclear, the study lays the groundwork for future research to explore how cells compensate for the absence of ORC and to develop functional approaches to investigate this process. The reviewers agree that this valuable paper would be strengthened significantly if the authors could delve a bit deeper into the nature of replication initiation, potentially using an origin mapping experiment. Such an exciting contribution would help explain the nature of the proposed new type of Mcm loading, thereby increasing the impact of this study for the field at large.

We appreciate the reviewers’ suggestion. Till now we know of only one paper that mapped origins of replication in regenerating mouse liver, and that was published two months back in Cell (PMID: 39293447). We want to adopt this method, but we do not need it to answer the question asked. We have mapped origins of replication in ORC-deleted cancer cell lines and compared to wild-type cells in Shibata et al., BioRXiv (PMID: 39554186) (it is under review). We report the following: Mapping of origins in cancer cell lines that are wild type or engineered to delete three of the subunits, ORC1, ORC2 or ORC5 shows that specific origins are still used and are mostly at the same sites in the genome as in wild type cells. Of the 30,197 origins in wild type cells (with ORC), only 2,466 (8%) are not used in any of the three ORC deleted cells and 18,319 (60%) are common between the four cell types. Despite the lack of ORC, excess MCM2-7 is still loaded at comparable rates in G1 phase to license reserve origins and is also repeatedly loaded in the same S phase to permit re-replication.

Citation: Specific origin selection and excess functional MCM2-7 loading in ORC-deficient cells. Yoshiyuki Shibata, Mihaela Peycheva, Etsuko Shibata, Daniel Malzl, Rushad Pavri, Anindya Dutta. bioRxiv 2024.10.30.621095; doi: https://doi.org/10.1101/2024.10.30.621095 (PMID: 39554186)

We have now included this in the discussion.

Public Reviews:

Reviewer #1 (Public review):

The origin recognition complex (ORC) is an essential loading factor for the replicative Mcm2-7 helicase complex. Despite ORC's critical role in DNA replication, there have been instances where the loss of specific ORC subunits has still seemingly supported DNA replication in cancer cells, endocycling hepatocytes, and Drosophila polyploid cells. Critically, all tested ORC subunits are essential for development and proliferation in normal cells. This presents a challenge, as conditional knockouts need to be generated, and a skeptic can always claim that there were limiting but sufficient ORC levels for helicase loading and replication in polyploid or transformed cells. That being said, the authors have consistently pushed the system to demonstrate replication in the absence or extreme depletion of ORC subunits.

Here, the authors generate conditional ORC2 mutants to counter a potential argument with prior conditional ORC1 mutants that Cdc6 may substitute for ORC1 function based on homology. They also generate a double ORC1 and ORC2 mutant, which is still capable of DNA replication in polyploid hepatocytes. While this manuscript provides significantly more support for the ability of select cells to replicate in the absence or near absence of select ORC subunits, it does not shed light on a potential mechanism.

The strengths of this manuscript are the mouse genetics and the generation of conditional alleles of ORC2 and the rigorous assessment of phenotypes resulting from limiting amounts of specific ORC subunits. It also builds on prior work with ORC1 to rule out Cdc6 complementing the loss of ORC1.

The weakness is that it is a very hard task to resolve the fundamental question of how much ORC is enough for replication in cancer cells or hepatocytes. Clearly, there is a marked reduction in specific ORC subunits that is sufficient to impact replication during development and in fibroblasts, but the devil's advocate can always claim minimal levels of ORC remaining in these specialized cells.

The significance of the work is that the authors keep improving their conditional alleles (and combining them), thus making it harder and harder (but not impossible) to invoke limiting but sufficient levels of ORC. This work lays the foundation for future functional screens to identify other factors that may modulate the response to the loss of ORC subunits.

This work will be of interest to the DNA replication, polyploidy, and genome stability communities.

Thank you.

Reviewer #2 (Public review):

This manuscript proposes that primary hepatocytes can replicate their DNA without the six-subunit ORC. This follows previous studies that examined mice that did not express ORC1 in the liver. In this study, the authors suppressed expression of ORC2 or ORC1 plus ORC2 in the liver.

Comments:

(1) I find the conclusion of the authors somewhat hard to accept. Biochemically, ORC without the ORC1 or ORC2 subunits cannot load the MCM helicase on DNA. The question arises whether the deletion in the ORC1 and ORC2 genes by Cre is not very tight, allowing some cells to replicate their DNA and allow the liver to develop, or whether the replication of DNA proceeds via non-canonical mechanisms, such as break-induced replication. The increase in the number of polyploid cells in the mice expressing Cre supports the first mechanism, because it is consistent with few cells retaining the capacity to replicate their DNA, at least for some time during development.

In our study, we used EYFP as a marker for Cre recombinase activity. ~98% of the hepatocytes in tissue sections and cells in culture express EYFP, suggesting that the majority of hepatocytes successfully expressed the Cre protein to delete the ORC1 or ORC2 genes. To assess deletion efficiency, we employed sensitive genotyping and Western blotting techniques to confirm the deletion of ORC1 and ORC2 in hepatocytes isolated from Alb-Cre mice. Results in Fig. 2C and Fig. 6D demonstrate the near-complete absence of ORC2 and ORC1 proteins, respectively, in these hepatocytes.

The mutant hepatocytes underwent at least 15–18 divisions during development. The inherited ORC1 or ORC2 protein present during the initial cell divisions, would be diluted to less than 1.5% of wild-type levels within six divisions, making it highly unlikely to support DNA replication, and yet we observe hepatocyte numbers that suggest there was robust cell division even after that point.

Furthermore, the EdU incorporation data confirm DNA synthesis in the absence of ORC1 and ORC2. Specifically, immunofluorescence showed that both in vitro and in vivo, EYFP-positive hepatocytes (indicating successful ORC1 and ORC2 deletion) incorporated EdU, demonstrating that DNA synthesis can occur without ORC1 and ORC2.

Finally, the Alb-ORC2f/f mice have 25-37.5% of the number of hepatocyte nuclei compared to WT mice (Table 2). If that many cells had an undeleted ORC2 gene, that would have shown up in the genotyping PCR and in the Western blots.

(2) Fig 1H shows that 5 days post infection, there is no visible expression of ORC2 in MEFs with the ORC2 flox allele. However, at 15 days post infection, some ORC2 is visible. The authors suggest that a small number of cells that retained expression of ORC2 were selected over the cells not expressing ORC2. Could a similar scenario also happen in vivo?

This would not explain the significant incorporation of EdU in hepatocytes that are EYFP positive and do not have detectable ORC by Western blots. Also note that for MEFs we are delivering the Cre by Adenovirus infection in vitro, so there is a finite probability that a cell will not receive the virus, the Cre and will not delete ORC2. However, in vivo, the Alb-Cre will be expressed in every cell that turns on albumin. There is no escaping the expression of Cre.

(3) Figs 2E-G shows decreased body weight, decreased liver weight and decreased liver to body weight in mice with recombination of the ORC2 flox allele. This means that DNA replication is compromised in the ALB-ORC2f/f mice.

It is possible that DNA replication is partially compromised or may slow down in the absence of ORC2. However, it is important to emphasize that livers with ORC2 deletion remain capable of DNA replication, so much so that liver function and life span are near normal. Therefore, some kind of DNA replication has to serve as a compensatory mechanism in the absence of ORC2 to maintain liver function and support regeneration.

(4) Figs 2I-K do not report the number of hepatocytes, but the percent of hepatocytes with different nuclear sizes. I suspect that the number of hepatocytes is lower in the ALB-ORC2f/f mice than in the ORC2f/f mice. Can the authors report the actual numbers?

We show in Table 2 that the Alb-Orc2f/f mice have about 25-37.5% of the hepatocytes compared to the WT mice.

(5) Figs 3B-G do not report the number of nuclei, but percentages, which are plotted separately for the ORC2-f/f and ALB-ORC2-f/f mice. Can the authors report the actual numbers?

In all the FACS experiments in Fig. 3B-G we collect data for a total of 10,000 nuclei (or cells). For Fig. 3E-G we divide the 10,000 nuclei into the bottom 40% on the EYFP axis (EYFP low, which is mostly EYFP negative) as the control group, and EYFP high (top 20% on the EYFP axis) test group. We have described this in the Methods in the revision and labeled EYFP negative and positive as EYFP low and high in the Figures and Figure legends.

(6) Fig 5 shows the response of ORC2f/f and ALB-ORC2f/f mice after partial hepatectomy. The percent of EdU+ nuclei in the ORC2-f/f (aka ALB-CRE-/-) mice in Fig 5H seems low. Based on other publications in the field it should be about 20-30%. Why is it so low here? The very low nuclear density in the ALB-ORC2-f/f mice (Fig 5F) and the large nuclei (Fig 5I) could indicate that cells fire too few origins, proceed through S phase very slowly and fail to divide.

The percentage of EdU+ nuclei in the ORC2f/f without Alb-Cre mice is 8%, while in PMID 10623657 ~10% of wild type nuclei incorporate EdU at 42 hr post partial hepatectomy (mid-point between the 36-48 hr post hepatectomy that was used in our study). The important result here is that in the ORC2f/f mice with Alb-Cre (+/-) we are seeing significant EdU incorporation. We have also corrected the X-axis labels in 5F, 5I, 7E and 7F to reflect that those measurements were not made at 36 hr post-resection but later (as was indicated in the schematic in Fig. 5A).

(7) Fig 6F shows that ALB-ORC1f/f-ORC2f/f mice have very severe phenotypes in terms of body weight and liver weight (about on third of wild-type!!). Fig 6H and 6I, the actual numbers should be presented, not percentages. The fact that there are EYFP negative cells, implies that CRE was not expressed in all hepatocytes.

The liver weight is very dependent on the body weight, and so we have to look at the liver to body weight ratio to determine if it is inordinately small, and the ratio is 70% of the WT. In females the liver and body weight are low (although in proportion to each other), which maybe is what the reviewer is talking about. However, the fact that liver weight and body weight are not as low in males, suggest that this is a gender (hormone?) specific effect and not a DNA replication defect. We had discussed this possibility. We have another paper also in BioRXiv (Su et al. doi.org/10.1101/2024.12.18.629220) that suggests that ORC subunits have significant effect on gene expression, so it is possible that that is what leads to this sexual dimorphism in phenotype. We have now added this to the discussion.

The bottom 40% of nuclei on the EYFP axis in the FACS profiles (what was labeled EYFP negative but will now be called EYFP low) contains mostly non-hepatocytes that are genuinely EYFP negative. Non-hepatocytes (bile duct cells, endothelial cells, Kupffer cells, blood cells) are a significant part of cells in the dissociated liver (as can be seen in the single cell sequencing results in PMID: 32690901). Their presence does not mean that hepatocytes are not expressing Cre. Hepatocytes are nearly 100% EYFP positive, as can be seen in the tissue sections (where the hepatocytes take up most of visual field) and in cells in culture. Also if there are EYFP negative hepatocyte nuclei in the FACS, that still does not rule out EYFP presence in the cytoplasm. The important point from the FACS is that the EYFP high nuclei (which have expressed Cre for the longest period) are polyploid relative to the EYFP low nuclei.

(8) Comparing the EdU+ cells in Fig 7G versus 5G shows very different number of EdU+ cells in the control animals. This means that one of these images is not representative. The higher fraction of EdU+ cells in the double-knockout could mean that the hepatocytes in the double-knockout take longer to complete DNA replication than the control hepatocytes. The control hepatocytes may have already completed DNA replication, which can explain why the fraction of EdU+ cells is so low in the controls. The authors may need to study mice at earlier time points after partial hepatectomy, i.e. sacrifice the mice at 30-32 hours, instead of 40-52 hours.

The apparent difference that the reviewer comments on stems from differences in nuclear density in the images in Fig. 7G and 5G (also quantitated in Fig. 7F and 5F). The quantitation in Fig. 7H and 5H show that the % of EdU plus cells are comparable (5-8%).

(9) Regarding the calculation of the number of cell divisions during development: the authors assume that all the hepatocytes in the adult liver are derived from hepatoblasts that express Alb. Is it possible to exclude the possibility that pre-hepatoblast cells that do not express Alb give rise to hepatocytes? For example the cells that give rise to hepatoblasts may proliferate more times than normal giving rise to a higher number of hepatoblasts than in wild-type mice.

Single cell sequencing of mouse liver at e11 shows hepatoblasts expressing hepatocyte specific markers (PMID: 32690901). All the cells annotated from the single-cell seq analysis are differentiated cells arguing against the possibility that undifferentiated endodermal cells (what the reviewer probably means by pre-hepatoblasts) exist at e11. We have added this citation to the paper.

Here is a review that says the hepatoblasts expressing Albumin are present before e13. (https://www.ncbi.nlm.nih.gov/books/NBK27068/) says: “The differentiation of bi-potential hepatoblasts into hepatocytes or BECs begins around e13 of mouse development. Initially hepatoblasts express genes associated with both adult hepatocytes (Hnf4α, Albumin) ...” Thus, we can be certain that hepatoblasts before e13 express albumin. Our calculation of number of cell divisions in Table 2 begins from e12.

The reviewer may be suggesting that ORC deletion leads to the immediate demise of hepatoblasts (despite having inherited ORC protein from the endodermal cells) causing undifferentiated endodermal cells to persist and proliferate much longer than in normal development. We consider it unlikely, but if true it will be very unexpected, both by suggesting that deletion of ORC immediately leads to the death of the hepatoblasts (despite a healthy reserve of inherited ORC protein) and by suggesting that there is a novel feedback mechanism from the death/depletion of hepatoblasts leading to the persistence and proliferation of undifferentiated endodermal cells. We have added the reviewer’s suggestion to the discussion.

(10) My interpretation of the data is that not all hepatocytes have the ORC1 and ORC2 genes deleted (eg EYFP-negative cells) and that these cells allow some proliferation in the livers of these mice.

Please see the reply in question #1. Particularly relevant: “Finally, the Alb-ORC2f/f mice have 25-37.5% of the number of hepatocyte nuclei compared to WT mice (Table 2). If that many cells had an undeleted ORC2 gene, that would have shown up in the genotyping PCR and in the Western blots.

Reviewer #3 (Public review):

Summary:

The authors address the role of ORC in DNA replication and that this protein complex is not essential for DNA replication in hepatocytes. They provide evidence that ORC subunit levels are substantially reduced in cells that have been induced to delete multiple exons of the corresponding ORC gene(s) in hepatocytes. They evaluate replication both in purified isolated hepatocytes and in mice after hepatectomy. In both cases, there is clear evidence that DNA replication does not decrease at a level that corresponds with the decrease in detectable ORC subunit and that endoreduplication is the primary type of replication observed. It remains possible that small amounts of residual ORC are responsible for the replication observed, although the authors provide arguments against this possibility. The mechanisms responsible for DNA replication in the absence of ORC are not examined.

Strengths:

The authors clearly show that there are dramatic reductions in the amount of the targeted ORC subunits in the cells that have been targeted for deletion. They also provide clear evidence that there is replication in a subset of these cells and that it is likely due to endoreduplication. Although there is no replication in MEFs derived from cells with the deletion, there is clearly DNA replication occurring in hepatocytes (both isolated in culture and in the context of the liver). Interestingly, the cells undergoing replication exhibit enlarged cell sizes and elevated ploidy indicating endoreduplication of the genome. These findings raise the interesting possibility that endoreduplication does not require ORC while normal replication does.

Weaknesses:

There are two significant weaknesses in this manuscript. The first is that although there is clearly robust reduction of the targeted ORC subunit, the authors cannot confirm that it is deleted in all cells. For example, the analysis in Fig. 4B would suggest that a substantial number of cells have not lost the targeted region of ORC2. Although the western blots show stronger effects, this type of analysis is notorious for non-linear response curves and no standards are provided. The second weakness is that there is no evaluation of the molecular nature of the replication observed. Are there changes in the amount of location of Mcm2-7 loading that is usually mediated by ORC? Does an associated change in Mcm2-7 loading lead to the endoreduplication observed? After numerous papers from this lab and others claiming that ORC is not required for eukaryotic DNA replication in a subset of cells, we still have no information about an alternative pathway that could explain this observation.

We do not see a significant deficit in MCM2-7 loading (amount and rate) in cancer cell lines where we have deleted ORC1, ORC2 or ORC5 genes separately in Shibata et al. bioRxiv 2024.10.30.621095; doi: https://doi.org/10.1101/2024.10.30.621095 (PMID: 39554186). This is now cited in the discussion.

The authors frequently use the presence of a Cre-dependent eYFP expression as evidence that the ORC1 or ORC2 genes have been deleted. Although likely the best visual marker for this, it is not demonstrated that the presence of eYFP ensures that ORC2 has been targeted by Cre. For example, based on the data in Fig. 4B, there seems to be a substantial percentage of ORC2 genes that have not been targeted while the authors report that 100% of the cells express eYFP.

(1) The PCR reactions in Fig. 4B are still contaminated by DNA from non-hepatocyte cells: bile duct cells, endothelial, Kupfer cells and blood cells. Microscopy of cultured cells idnetifies the hepatocytes unequivocally from their morphology. <2% of the hepatocyte cells in culture in Fig. 4C are EYFP-.

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

The authors should present the data as suggested in the review and reformulate their conclusions. If possible, mice should be examined 30-32 hours after partial hepatectomy.

Based on the Literature we chose a time that is consistent with the previous paper from us (Uchida et al., Genes & Dev).

Reviewer #3 (Recommendations for the authors):

(1) It would improve the paper to use single-cell methods (e.g. FISH) to assess the deletion of ORC subunits in the targeted cells.

This is something we will reserve for future studies.

(2) The importance of the paper would be increased dramatically by showing that the elimination of ORC changed the location of Mcm2-7 loading. This would be highly likely if the authors hypothesis that ORC is not involved is true. On the other hand, given ORC's role in origin selection, an observation that the same sites are used but less frequently would support a hypothesis that residual intact ORC is responsible for the replication observed.

Shibata et al (PMID: 39554186) has answered this question. The loss of ORC does not change the locations of origins or even the ability to specify origins. We argue that this is what is to be expected from our hypothesis, that although ORC is clearly important for MCM loading in yeast and in biochemical experiments, something unexpected is going on in human cells. Either a vanishingly small amount of ORC (undetectable by commonly used methods) can load the full complement of MCM2-7 at a rate that is comparable to wild type cells, or there is an ORC-independent mechanism of MCM2-7 loading. This is now added to the discussion.

Version 2

Abstract

Abstract

The six subunit ORC is essential for initiation of DNA replication in eukaryotes. Cancer cell-lines in culture can survive and replicate DNA replication after genetic inactivation of individual ORC subunits, ORC1, ORC2 or ORC5. In primary cells, ORC1 was dispensable in the mouse liver for endo-reduplication, but this could be explained by the ORC1 homolog, CDC6, substituting for ORC1 to restore functional ORC. Here, we have created mice with a conditional deletion of ORC2, which does not have a homolog. Although mouse embryo fibroblasts require ORC2 for proliferation, mouse hepatocytes synthesize DNA in cell culture and endo-reduplicate in vivo without ORC2. Mouse livers endo-reduplicate after simultaneous deletion of ORC1 and ORC2 both during normal development and after partial hepatectomy. Since endo-reduplication initiates DNA synthesis like normal S phase replication these results unequivocally indicate that primary cells, like cancer cell lines, can load MCM2-7 and initiate replication without ORC.

Introduction

DNA replication during normal mitotic cycles or during endo-reduplication is initiated by the six subunit ORC, five of which have AAA+ATPase domains and form a ring shaped complex that binds DNA and bends it (Bell and Stillman, 1992, Neuwald et al., 1999, Dhar et al., 2001a, Clarey et al., 2006, Bleichert et al., 2015). In cooperation with another ATPase, CDC6, and with CDT1, ORC helps load MCM2-7 double hexamers at or near origins of replication(Costa and Diffley, 2022, Stillman, 2022, Hu and Stillman, 2023). To initiate replication (or endo-reduplication), the MCM2-7 hexamer associates with additional proteins to form the functional CMG helicase that is essential for unwinding the double stranded chromosomal DNA, so that the resulting single-stranded DNAs can serve as templates for copying by DNA polymerases. In the lower eukaryotes, various ORC subunits have been consistently found to be essential for DNA replication and cell proliferation (Bell et al., 1993, Foss et al., 1993, Micklem et al., 1993, Loo et al., 1995, Semple et al., 2006). In mammalian cells, knockdown of individual ORC subunits have also been reported to stop cell proliferation (Chou et al., 2021, Prasanth et al., 2004). In humans, mutations in Orc1, Orc4, Orc6, Cdt1, and Cdc6 cause microcephalic primordial dwarfism resembling Meier-Gorlin syndrome (Bicknell et al., 2011b, Bicknell et al., 2011a, Guernsey et al., 2011).

There have, however, been isolated reports that suggest that under unusual circumstances, DNA replication can initiate in eukaryotes in the absence of the full ORC. First hypomorphic mutation made in the Orc2 gene in HCT116 colon cancer cells decreased ORC2 protein levels by 90% and cells were still viable and could proliferate (Dhar et al., 2001b). Deletion of the Orc1 gene in Drosophila permitted several rounds of replication in the resulting larvae and pupae (Park and Asano, 2008). A few human cancer cell lines have been created by CRISPR-Cas9 mediated genome engineering where ORC1, ORC2 and ORC5 cannot be detected by regular immunoblots and yet the cell lines survive, proliferate and replicate DNA with the normal complement of origins of replication (Shibata et al., 2016, Shibata and Dutta, 2020). However, the Orc2 gene is essential for viability and in the Orc2Δ cell line a very minimal level (<0.1% of wild type levels) of a truncated protein can be detected that reacts with anti-ORC2 antibody and co-immunoprecipitates with ORC3 (Chou et al., 2021). Given that WT cells have about 150,000 molecules of ORC2, even if this truncated protein is functional ORC2, ∼150 molecules of the protein would be expected to load MCM2-7 double hexamers on at least 50,000 origins of replication. This possibility would be consistent with current models of MCM2-7 loading only if ORC was highly catalytic and one ORC hexamer was capable of loading ∼667 MCM2-7 hexamers (Costa and Diffley, 2022, Stillman, 2022, Hu and Stillman, 2023).

To address whether the entire ORC holocomplex is essential for DNA replication initiation in primary mammalian cells, we have reported a conditional mutation in the ORC1 gene of mouse (Okano-Uchida et al., 2018). Conditional deletion of Orc1 (using tissue-specific Cre drivers) revealed that the gene was essential for normal mitotic cell division in intestinal epithelial cells, but that endoreduplication in tissues that are known to become polyploid, like placental trophoblasts and the liver hepatocytes, was unaffected. Because the ORC1 protein is a functional ATPase with sequence homology with CDC6 (Saha et al., 1998, Williams et al., 1997), one possibility that remained was that the mouse CDC6 protein substituted for ORC1 in the ORC holocomplex, producing enough functional ORC. Therefore, we decided to make a conditional mutation in another subunit of ORC, Orc2, that does not have the Walker A or B motifs that would make it a functional ATPase and that is not homologous to Cdc6. We designed the mutation such that even if an N terminally truncated protein is expressed from a downstream methionine from an alternately spliced mRNA the AAA+ like domain will suffer a significant deletion.

This will allow us to test whether the mouse liver survival and endoreduplication after loss of ORC1 is seen even after loss of ORC2. It will also test whether combining the conditional mutations of ORC1 and ORC2 will allow the liver cells to survive and endoreduplicate. If hepatocyte survival and endoreduplication is seen in these mice, then it would be hard to explain this by the substitution of two ORC subunits by the CDC6 protein, and we would gain support for the hypothesis that in rare circumstances even untransformed mammalian cells are capable of loading enough MCM2-7 in the absence of ORC to permit extensive DNA replication.

Results

ORC2 is essential for embryonic development and proliferation of mouse embryo fibroblasts

Mice with Lox P sites inserted flanking exons 6 and 7 of mouse Orc2 were purchased from Cyagen Biosciences Inc. (Fig. 1A). Genotyping with primers F2 and R2 distinguished the loxP marked allele from the WT allele (Fig. 1B). In a cross of Orc2f/+ mice, there was no significant decrease in the yield of Orc2f/+ or Orc2f/f compared to Orc2+/+ mice (Fig. 1C), suggesting that the insertion of the loxP sites in the introns of Orc2 did not impair the function of ORC2. Recombination between the lox2 sites would delete the coding exons 6 and 7 of Orc2, which removes amino acids L111-V150 (Fig. 1D). If a transcript is expressed that skips exons 6 and 7, then the resulting protein mutates A110 to V110 and then throws the protein coding sequence out of frame (Fig 1D, ΔORC2), so that the bulk of the 576 amino acid ORC2 protein, including the AAA+ like and WH domains (K230-A576) are not expressed. The AAA+ like domain and the WH domain are key elements for ORC2 assembly into ORC and for ORC function. Even in the remote possibility that a truncated protein is expressed due to alternative splicing and translation initiation from an internal methionine, the next methionine is at M301, so that half of the protein including 70 amino acids of the 239 amino acid AAA+-like domain (K230-G469) would be deleted.

Generation of Orc2f/fmice and ORC2 KO MEFs.

A. Scheme of introduced loxP sites in Orc2 locus.

B. Representative picture of genotyping of offspring coming from Orc2f/+ crossed with Orc2f/+.

C. The ratio of observed to expected animals coming from Orc2f/+ crossed with Orc2f/+.

D. Schematic of the ORC2 protein and the DeltaORC2 protein produced after deletion of exons 6 and 7. A110 is mutated to V110 and then the protein goes out of frame.

E. Validation of Orc2 deletion 3 days after Adeno cre transduction.

F. Western blot of ORC2 protein 5 days after Adeno cre transduction. 10 or indicated ul of lysate loaded/lane as written on the top

G. MTT assay of WT and Orc2f/f MEFs without and with Adeno cre transduction.

H. Western blot of ORC2 protein 5 and 15 days after Adeno cre transduction.

The Sox2-Cre allele expresses active Cre during embryonic development. Crossing the Orc2f/f mice with Sox2-Cre mice resulted in no Orc2Δ/Δ embryos at E7.5 days onwards suggesting that embryonic deletion of Orc2 is lethal (Table1). The near expected number of Orc2+/Δ embryos suggests that hemizygosity of Orc2 can still support viability.

Mouse embryo fibroblasts (MEF) were obtained from Orc2f/f E13.5 day embryos and cultured in vitro. 3 days after infection with an Adeno virus expressing Cre, effective recombination between the loxP sites is seen, resulting in a genotype showing that exons 6 and 7 have been deleted in most of the MEF (Orc2Δ) (Fig. 1E). Consistent with this, at 5 days after adeno-Cre infection, the ORC2 protein is not detected in the MEF population (Fig. 1F). The dilution of the undeleted MEF lysate on the same blot suggests, that if any ORC2 protein is expressed in the MEF after adeno-Cre infection, it is <10% of that in the undeleted MEF. The proliferation rate of the MEF were measured by MTT assays. Even in the Orc2+/+ MEF, the infection with adeno-Cre decreased proliferation a little (the orange line compared to the blue line in Fig. 1G). However, for Orc2f/f MEF infection with adeno-Cre impairs proliferation even further (yellow line compared to black line in Fig. 1G). Furthermore, when the MEF were cultured for 15 days, the surviving cells express significant levels of ORC2 (Fig. 1H), suggesting that the MEFs that had not undergone the Cre-mediated deletion take over the culture.

Taken together, these results suggest that Cre mediated deletion of exons 6 and 7 of Orc2 leads to early embryonic lethality and impairs the proliferation of normal diploid MEFs in culture.

Knockout of Orc2 in developing mouse liver makes ORC2 protein undetectable in hepatocytes and yet supports most of normal development and endoreduplication

Mice carrying one copy of the Albumin promoter driven Cre gene (Alb-Cre) express the Cre recombinase specifically in hepatocytes (Postic et al., 1999). When the Orc2f/f mice are crossed with mice carrying Alb-Cre (shortened below as Alb), the Cre recombinase is expected to promote the recombination mediated deletion of exons 6 and 7 of Orc2 in the hepatocytes. The Alb+/--Orc2f/f mice were viable and fertile (Fig. 2A, 2B). Crossing such mice yielded the expected numbers of Alb+/--Orc2f/f mice, though there was a drop in the yield of Alb+/+-Orc2f/f mice. One allele of Alb-Cre is sufficient to express enough Cre to carry out homozygous deletion of the Orc2 allele in the hepatocytes, so that the results suggest that deletion of Orc2 in the hepatocytes did not impair viability. The partial lethality of the Alb+/+ (HOM) mice could be due to the toxicity of high dose of Cre recombinase expressed, something that has been noted by other groups (Schmidt et al., 2000, Loonstra et al., 2001, Schmidt-Supprian and Rajewsky, 2007, Naiche and Papaioannou, 2007, Janbandhu et al., 2014). Because the Alb-Cre is expressed only in the hepatocytes and not in other cells in the liver, we isolated hepatocytes from these livers by growing them in culture (schematically shown in Fig. 4A) to determine whether the ORC2 protein was decreased (Fig. 2C). The ORC2 protein was significantly decreased in all five of the Alb+/-- Orc2f/f mice, showing consistent and near 100% effect of the deletion in the hepatocytes. Western blots for several other ORC subunits, and CDC6 protein showed that the loss of ORC in the same hepatocytes did not significantly decrease these proteins, though MCM2 and MCM3 (two subunits of the MCM2-7 helicase component) were decreased by 50% (Fig. 2D). The body weights of the Alb, Orc2f/f mice were smaller than in the Orc2f/f animals of both sexes, though the liver size and liver size normalized to body weight was significantly smaller only in the females (Fig. 2E-G). There was some elevation of the circulating liver enzymes in the mice where Alb-Cre is expressed to delete the Orc2 gene, suggesting that there is some impairment of liver function (Supp. Fig. S1).

Development of liver in Orc2 KO mice.

A. Scheme of Alb+/--Orc2f/f ROSA26stop-EYFP crossed with Alb+/--Orc2f/f ROSA26stop-EYFP (All mice are with ROSA26stop-EYFP and so we do not include this in the genotypes below)

B. The ratio of observed to expected animals coming from A.

C. Western blot of hepatocytes from Orc2 f/f and Orc2 f/f Alb-Cre +/- animals. Tubulin was used as loading control.

D. Quantification of the Western blots of hepatocyte lysates from Orc2 f/f (without Alb-cre) mice and the same genotype but with Alb-Cre to show the levels of other key replication initiation proteins in the ORC2 KO hepatocytes.

E. Average body weight of Orc2f/f and Alb-Orc2f/f animals.

F. Average liver weight of Orc2f/f and Alb-Orc2f/f animals.

G. Average liver to body weight ratio of Orc2f/f and Alb-Orc2f/f animals.

H. Representative H&E staining of liver tissue from Orc2f/f (WT) and Alb-Orc2f/f (KO) animals. Both panels at same scale.

I. Quantification of hepatocyte nuclear size in Orc2f/f and Alb-Orc2f/f animals.

J. Quantification of hepatocyte nuclear size in Orc2f/f and Alb-Orc2f/f female mice.

K. Quantification of hepatocytes nuclear size in Orc2f/f and Alb-Orc2f/f male mice.

*p < 0.05, **p < 0.01, two-tailed Student’s t test.

The livers of mice can grow by proliferation (increasing the number of cells) or by hypertrophy of cells (larger cells with endoreduplicated nuclei). Histological examination of the livers revealed that the Orc2 deletion was accompanied by the presence of fewer (∼50% of WT levels), but larger nuclei and cells (Fig. 2H and 2I). The Orc2 deleted hepatocytes had significantly larger nuclei in both males and females (Fig. 2J, K). The larger nuclei in hepatocytes with Orc2 deletion also seen in other contexts, e.g. in the pre-resection hepatocytes with Orc2 deletion (Fig. 5E, 0 hr point) and in the Cre expressing, Orc1f/+, Orc2f/f that have an intact Orc1 gene but delete both alleles of Orc2 (Supp. Fig. S3B, green bars compared to blue bars).

To determine whether the larger nuclei were generated by endoreduplication, we isolated nuclei from hepatocytes and stained them with the DNA staining dye, DRAQ5, followed by flow cytometry (Fig. 3A). Both male and female Alb-Cre, Orc2f/f hepatocytes showed a decrease in 4n DNA content and an increase in 16n DNA content (Fig. 3B-D). The expression of Cre would also delete a lox-stop-lox element upstream from EYFP and lead to the expression of EYFP. Indeed, when we gated on EYFP positive cells, the 8n and 16N DNA content increase was much more evident relative to the EYFP negative cells, while the 2N DNA containing nuclei were decreased (Fig. 3E-G). Thus Orc2 deletion promotes endo-reduplication, a result similar to what was noted when Orc1 was deleted in mouse livers (Okano-Uchida et al., 2018).

Endoreduplication in the Orc2 liver conditional knock-out animals.

A. Experimental design.

B-D. Quantification of nuclei ploidy on 10,000 nuclei from the livers of Orc2f/f ROSA26stop-EYFP and Alb-Orc2f/f ROSA26stop-EYFP animals.

E-G. Quantification of nuclei ploidy for EYFP low (includes negative) and high (positive) primary liver cells. *p < 0.05, **p < 0.01, ***p < 0.001, two-tailed Student’s t test.

DNA replication of Alb-Cre+/--Orc2f/f hepatocytes in vitro

Primary hepatocytes isolated from 8-10 weeks old mouse liver can replicate their DNA and proliferate for a limited time in vitro. We isolated such hepatocytes from Alb-cre-/- (WT) and Alb-cre+/- livers, both from Orc2f/f mice (Fig. 4A). Genotyping reveals that over 90% of the cells from the Alb-cre+/- livers have successfully deleted Orc2 exons 6 and 7 (Fig. 4B), while immunoblotting shows that ORC2 protein expression is also significantly decreased in these cells (Fig. 3C). EdU labeling in vitro for 2 hrs, showed that the EYFP positive cells (where Cre recombinase has been active) can incorporate EdU (Fig. 4C), although the total number of nuclei that incorporate EdU is decreased to about 30% of that seen in the ORC2 wild type hepatocytes (Fig. 4D). Thus hepatocytes in culture can continue to replicate DNA in the absence of detectable ORC2 protein.

Orc2 KO primary hepatocytes are viable and can incorporate EdU in vitro.

A. Experimental design

B. Genotyping and western blotting of hepatocytes

C. Representative picture of EdU, EYFP and DAPI staining on the Orc2 WT (Orc2f/f) and KO (Orc2f/f Alb-Cre) primary hepatocytes.

D. The percentage of EdU positive nuclei from Orc2 WT or Orc2 KO primary hepatocytes.

*p < 0.05, two-tailed Student’s t test.

DNA replication in Alb-Cre+/--Orc2f/f hepatocytes during liver regeneration in vivo

Liver DNA synthesis peaks around 36-48 hours, and the liver regenerates to nearly 50% of its original weight within 2-3 days following partial hepatectomy. Such regeneration involves both normal mitotic DNA replication/cell division and endoreduplication accompanied by hypertrophy of the cells. We therefore tested whether deletion of the Orc2 gene in the hepatocytes adversely affects liver regeneration after partial hepatectomy in 8-14 weeks old mice (Fig. 5A). Nearly two-thirds of the liver is removed surgically and the mice allowed to recover for 36-48 hrs before harvesting the regenerated liver.

The ORC2 mutant livers regenerate after partial hepatectomy.

A. Schematic of the experiment.

B. Body weight of the Orc2f/f ROSA26stop-EYFP mice without (-/-) or with Alb-Cre (-/+) before partial hepatectomy.

C. Liver weight of the mice in B after liver regeneration.

D. Regenerated Liver to pre-hepatectomy body weight ratio of the mice in B.

E. H&E stain of Orc2f/f ROSA26stop-EYFP livers with intact Orc2 (Alb-cre -/-, N=3) or Orc2 knockout (Alb-cre+/-, N=7). Scale bar: 25 μm.

F. Quantitation of nuclear counts per field (76,000 um2). Six images were taken for each liver. 0 hr (pre- resection). 42 hr (post-regeneration).

G. EdU incorporation of indicated livers. EYFP marks cells where Cre has been expressed. Scale bar: 25 μm.

H. Percent EdU+ nuclei counted in 1882 and 825 nuclei in the Cre- and Cre+ livers, respectively.

I. Nuclear size of indicated livers. 0 hr (pre-resection). 42 hr (post-regeneration). Mean and S.D from about 40-70 nuclei, *p < 0.05, ****p < 0.0001, unpaired two-tailed Student’s t test is used

Although the liver weight after regeneration was smaller than that of WT livers, the liver:body weight ratio was similar in the livers with Orc2 deletion and Orc2 WT (Fig. 5B-D). The H&E stain of the regenerated liver shows that the liver cells are larger and have larger nuclei in the liver with Orc2 deletion (Fig. 5E). In these experiments, both before and after regeneration, the livers with the Orc2 deletion have ∼50% nuclei compared to the wild type livers (Fig. 5F). Finally, labeling the livers by injection of EdU in the mice 3-4 hrs before harvesting, shows that the EYFP positive cells (indicating activity of Cre recombinase) were proficient in synthesizing DNA and incorporating EdU (Fig. 5G, H).

Nearly 100% of the hepatocytes were positive for EYFP, suggesting that the Cre recombinase was active in the vast majority of the hepatocytes (Fig. 5G), and consistent with the complete depletion of Orc2 protein from the hepatocytes of all mice with that genotype (Fig. 3C). Thus the 30% of the liver cells that were incorporating EdU in the Cre expressing livers (Fig. 5H) were doing so in the absence of Orc2 protein. Also, we noted several pairs of EdU positive nuclei (marked by arrows in Fig. 5G) whose relative positions suggest that they are sisters born from the same mitosis, suggesting that some EdU positive cells can go through mitosis in the EYFP positive cells. Finally, the nuclei were significantly larger in the Cre active Orc2f/f hepatocytes than in the hepatocytes without Cre, both before and after regeneration (Fig. 5I).

We did not explore why the EYFP protein is mostly nuclear in hepatocytes in culture (Fig. 4C) and mostly cytoplasmic in hepatocytes in the liver tissue (Fig. 5G, 7G), but speculate that differences in signaling pathways or fixation techniques between the two conditions contribute to this difference.

Viable mice with endoreduplicated hepatocyte nuclei in Alb-Orc1f/f Orc2f/f mice

We have reported that conditional deletion of Orc1 in developing mouse livers still allowed livers to develop and induced premature endoreduplication, suggesting that significant DNA synthesis can occur in liver cells that are genetically deleted of Orc1. We bred the Orc1f/f mice with Orc2f/f mice to obtain mice where both Orc1 and Orc2 are floxed but Alb-cre is not present (Fig. 6A, B). Even though Cre was not expressed in these livers, for unknown reasons there was a decrease in the percentage of progeny when Orc2 was floxed in the liver (with or without Orc1 being floxed), but not when just Orc1 was floxed (Fig. 6C). These mice were then bred with Orc2f/f Alb-cre+/+ mice, and the resulting Orc1f/+, Orc2f/f, Alb-cre+/- mice intercrossed to get Orc1f/f, Orc2f/f, Alb-Cre+/- mice where the Orc1 and Orc2 alleles are conditionally deleted in hepatocytes, and the deletion is accompanied by activation of EYFP expression. Immunoblotting of isolated hepatocytes showed that in four out of four mice expressing the Cre recombinase the ORC1 and ORC2 proteins were decreased significantly (Fig. 6D).

Endoreduplication in liver of Orc1 Orc2 DKO animals.

A-B. Breeding schemes to obtain conditional double flox animals.

C. The ratio of observed to expected animals coming from B. Orc1 = all animals with Orc1f/f ROSA26stop-EYFP, Orc2 = all animals with Orc2f/f ROSA26stop-EYFP, Orc1 Orc2 = all animals with Orc1f/f Orc2f/f ROSA26stop-EYFP genotype. This was before introduction of Alb-Cre.

D. Immunoblot of hepatocytes from WT (Orc1f/f Orc2f/f) and DKO (Orc1f/f Orc2f/f Alb-cre+/-) mice to show that ORC1 and ORC2 are depleted in the DKO cells.

E. Quantitation of immunoblots to show that levels of other key initiation protein subunits are not decreased in the DKO mice hepatocytes.

F. Average body, liver weight and their ratio for WT and DKO animals.

G. Representative H&E staining of liver tissue from male WT and DKO animals.

H. Quantification of hepatocyte nuclear size in the WT and DKO animals.

I. Quantification of nuclei ploidy for EYFP low (includes negative) and high (positive) primary liver cells from DKO mice.

Immunoblotting of some of the other replication proteins showed no decrease of other ORC subunits, CDC6 and two of the MCM2-7 subunits (Fig. 6E). The ORC3 protein, which did not change when Orc2 was deleted (Fig. 2D) was surprisingly elevated in Fig. 6E when Orc1 and Orc2 were both deleted. We do not know why this is the case, and it was not seen consistently in all four animals. Similarly, the 50% decrease of MCM2 and MCM3 that was seen in the absence of ORC2 (Fig. 2D), disappeared when both Orc1 and Orc2 are deleted.

The female Alb-Orc1f/f Orc2f/f mice were significantly smaller in size, with smaller livers and decreased liver to body weight ratio (Fig. 6F, Supp. Fig. 2A), but the male mice were relatively less affected (Fig. 6F). We have also observed ∼50% lethality of double knock-out female mice within the first month of life, but the remaining 50% survive beyond 4 months of age (Supp. Fig. 2C). H&E staining showed that the double knockout (DKO) male livers have fewer but larger cells with significantly larger nuclei (Fig. 6G, 6H). Interestingly, this phenotype is even more marked in the female liver (Supp. Fig. 2B). By flow cytometry the EYFP positive cells as evidence of Cre recombinase activity had distinctly more polyploidization than the EYFP negative cells (Fig. 6I). Taken together the results reveal that hepatocytes can synthesize DNA by endoreduplication to produce very large nuclei and very large cells so that the liver size is not hugely decreased, even in the absence of two subunits of ORC. Female mice that have a deletion in two subunits of ORC suffer more morbidity and mortality than male mice.

DNA replication in Alb-Cre+/--Orc1f/f Orc2f/f hepatocytes during liver regeneration in vivo

Finally, partial hepatectomy was performed in the livers of male mice with DKO of Orc1 and Orc2. As can be seen even before partial hepatectomy the body weights were not significantly different from WT mice (Fig. 7A). Post regeneration, the liver weights and liver/body weight ratios were not decreased in the DKO mice compared to the WT mice (Fig. 7B, C). H&E sections showed that the hepatocyte nuclei were larger and hepatocyte nuclear density lower in the DKO livers both pre- hepatectomy and in the regenerated livers 36 hr post-hepatectomy (Fig. 7D-F). EdU labeling for 3-4 hrs before harvesting the livers showed that despite the DKO, the EYFP positive cells showed extensive DNA synthesis (Fig. 7G). Even though nearly 100% of the hepatocytes are EYFP positive (Fig. 7G) and ORC1 and ORC2 proteins depleted in the hepatocytes (Fig. 6D), 15 % of nuclei stained with EdU (Fig. 7H), a percentage that was higher in the DKO regenerating livers than in the WT livers, suggesting that the endoreduplication that accompanies liver regeneration can occur even after deletion of two of the six subunits of ORC.

Endoreduplication in the liver of the DKO mice during liver regeneration.

A-C. Body weight pre-resection, liver weight post-regeneration and regenerated liver to body weight ratio in mice with indicated genotypes. Black bars: 4 WT males (Orc1f/f Orc2f/f ROSA26stop-EYFP mice without Alb-Cre). White bars: 6 DKO males (Orc1f/f Orc2f/f ROSA26stop-EYFP mice with Alb-Cre +/-). No significant difference between the two groups using two tailed Student t-test.

D: H&E stain of WT (N=4) or DKO mice (N=6). Scale bar: 50 μm.

E. Quantitation of hepatocyte nuclear size post regeneration. WT: black bars. DKO: white bars. 0 hr (pre-resection). 42 hr (post-regeneration). Five-six images were taken for each liver. About 120-200 nuclei are counted.

F: Quantitation of hepatocyte nuclear density post regeneration. WT: black bars. DKO: white bars. 0 hr (pre-resection). 42 hr (post-regeneration). Five-six images were taken for each liver.

G. Micrographs of EdU, DAPI and EYFP imaging of livers with indicated genotypes post regeneration. Scale bar: 20 μm. WT in the top row, DKO in the bottom row.

H. Quantitation of EdU positive nuclei post regeneration. WT: black bar. DKO: white bar. Five-six images were taken for each liver. *p < 0.05, ****p < 0.0001, unpaired two-tailed Student’s t test were used

Discussion

The six subunit ORC is important for recruiting MCM2-7, the core of the replicative DNA helicase and is essential for all forms of DNA replication in eukaryotic cells(Costa and Diffley, 2022, Stillman, 2022, Hu and Stillman, 2023). Yet, it has been possible to select cancer cell-lines that have mutations in Orc1, Orc2 or Orc5 and do not express detectable levels of the proteins and yet load MCM2-7 to the chromatin and replicate their DNA as they proliferate in culture (Shibata et al., 2016, Shibata and Dutta, 2020). This paradox suggested that there may be an alternative way to load sufficient MCM2-7 on chromatin and support DNA replication, at least in cancer cells in culture. In addition, there have been two instances where Orc1 has been mutationally removed in Drosophila (Park and Asano, 2008) and in mouse hepatocytes or placental trophoblasts (Okano-Uchida et al., 2018), where extensive DNA replication (particularly endo-reduplication in hepatocytes and trophoblasts) has persisted, suggesting that an alternative way of loading MCM2-7 may be available in certain unique cell-cycles that are different from the normal DNA replication that occurs during normal mitotic growth of diploid cells.

Five of the six subunits of ORC (ORC1-5) associate to form a ring-shaped structure with interactions between the WH domains and the AAA+ like domains of the five subunits (Bleichert et al., 2015). In mammalian cells only ORC1 and ORC4 have the intact Walker A and B motifs that are necessary for the molecules to act as AAA+ ATPases (Giordano-Coltart et al., 2005). Though this has never been demonstrated experimentally, it is theoretically possible that in the absence of ORC1 in flies or in mouse livers, another related AAA+ ATPase, intimately involved in MCM2-7 loading, CDC6, can substitute for ORC1 to reconstitute a functional 5 subunit ring-shaped ORC-like structure that executes its function (Takeda et al., 2005, Bell, 2017). We therefore wished to test whether removal of a second subunit of ORC, ORC2, that does not have obvious ATPase activity and not much homology to CDC6, would still permit endoreduplication in mouse livers. In the cancer cells, we had removed the initiator methionine of the Orc2 gene (Shibata et al., 2016). Although 99.9% of the ORC2 protein disappeared, an N terminally truncated form of ORC2 protein was expressed in the mutant cells at 0.1% of the wild type level (Chou et al., 2021). Thus, in this case we took the additional precaution to design a mutation such that even if a truncated protein was expressed from an internal methionine downstream from the mutation site, such a protein will be missing a significant part of its AAA+ATPase domain. Anyhow, in Western blots we do not see any evidence of such a truncated protein being expressed in the genetically altered MEFs or hepatocytes. In addition, we wondered whether the simultaneous genetic removal of two subunits of ORC, Orc1 and Orc2, would successfully obliterate all forms of DNA replication, including endoreduplication.

Our results show that while Orc2 is genetically essential for viability of early embryos in utero and mouse embryo fibroblasts in vitro, mutational inactivation of Orc2, or combined inactivation of Orc1 and Orc2, does not significantly inhibit development of the mouse liver. Adult, viable mice are produced. Albumin expression is activated very early in hepatoblasts in the 7-8 somite stage and the albumin mRNA can be seen in the hepatic primordium emerging from the gut at E9.5 days (Gualdi et al., 1996). Single cell sequencing studies confirm that hepatoblasts expressing albumin are present at e11 and that undifferentiated endodermal cells (the precursors to hepatoblasts) are non-existent (Wang et al., 2020) 3H-thymidine incorporation can be measured in the liver as late as days 7-14 postnatally (Tilghman and Belayew, 1982), so that many cycles of DNA replication and cell proliferation are expected to occur in hepatoblasts and hepatocytes after the activation of Alb-Cre and subsequent homozygous knockout of Orc2 (or Orc1) in the hepatocytes.

The presence of functional livers in the mutant (but viable) animals, suggested that the homozygous knockout of these genes did not have a profound effect on liver development, as would have been expected if the cells were as sensitive as MEFs in culture to the loss of the ORC holocomplex. We have been unable to find antibodies that will recognize mouse Orc1 or Orc2 proteins in immunohistochemistry on tissues and so decided to calculate how many cell divisions have to occur after the Orc1 or Orc2 genes are deleted in the embryonic mouse hepatocytes (see Methods and Table 2). The calculations suggest that the Orc2 deleted livers and male DKO livers underwent at least 18 cell divisions, while the female DKO livers underwent at least 15 cell divisions since E9.5 (Table2, bottom row).

If we allow a generous 3 cell divisions for the Cre recombinase to definitively excise the Floxed genes (consistent with what we see in MEFs in culture), this means that 15 cell divisions in the Orc2 deleted livers and male DKO livers occurred after the gene(s) was/were deleted. Once a gene is deleted, each cell division decreases the corresponding protein at least by half, so that 6 cell divisions would dilute the targeted protein to <1.5% of the wild type levels. Thus even after the ORC2 (or ORC1 and ORC2 in the DKO mice) decreased to <1.5% the WT level, the Orc2 knockout livers (male or female) and the male DKO livers could execute at least 9 more rounds of replication and cell division. Note the amount of the relevant ORC subunit would continue to decrease by half with each further division. Also, that the levels of ORC are likely to decrease faster if the protein is actively degraded or if more cell divisions are necessary to counter any development-related apoptosis.

This suggests that mitotic DNA replication of diploid cells is not completely eliminated at least in the Orc2-/- or male DKO livers after loss of the targeted ORC proteins. Consistent with this, we see EdU incorporation in EYFP+ hepatocytes, in vitro, which, however, could be from a terminal endo-reduplication event. More tantalizing is the presence of paired EYFP positive EdU-positive nuclei during the regeneration after partial hepatectomy in 8-14 week old mice (Fig. 5G). Such paired nuclei are usually seen with daughter nuclei that have just been born from mitosis, suggesting that there were at least a few cells that underwent complete DNA replication and mitosis even though the hepatocytes had undergone 15 rounds of cell division after deletion of the targeted ORC gene(s).

Could the ORC deletion lead to the immediate loss of hepatoblasts (despite having inherited ORC protein from the endodermal cells) causing undifferentiated endodermal cells to persist and proliferate much longer (without activating Cre) than in normal development? We consider this unlikely, but if true it will be very unexpected, both by suggesting that deletion of ORC immediately leads to the death of the hepatoblasts (despite a healthy reserve of inherited ORC protein) and by suggesting that there is a novel feedback mechanism from the death/ORC depletion of hepatoblasts leading to the persistence and proliferation of undifferentiated endodermal cells.

The female DKO (but not the Orc2-/-) livers seem more sensitive to the depletion of both ORC1 and ORC2 since ∼12 cell divisions in the female DKO livers occur after allowing 3 cell division for the targeted genes to be definitively deleted. Thus, after the cells have reached <1.5 % of WT levels of ORC2 or ORC1 protein, they can still execute at least 6 more cell divisions. Consistent with this sex specific difference, the male mice appear to be more tolerant of deletion of Orc2 gene or both Orc1 and Orc2 genes, with the livers reaching larger sizes than in the female mice. Plasma testosterone levels reach 50% of adult levels in the first day after birth before declining (to rise again at puberty at 4-5 weeks age), suggesting that enough androgens are present during the proliferative phase of liver development (Bell, 2018). Sex-specific changes in gene expression have been noted to begin in the liver by 3 weeks postnatally (Conforto and Waxman, 2012). Thus, it is possible that androgens directly or indirectly stimulate the division or hypertrophy of hepatocytes to ameliorate the negative effects on liver mass due to the loss of ORC. We have also shown that loss of ORC subunits in cancer cell lines causes significant changes in gene expression due to the role of ORC subunits in epigenetic regulation(Su et al., 2024), so another explanation of the more severe phenotype in females could be that the epigenetic effects of ORC subunits are more important in female compared to male livers.

In contrast to diploid DNA replication in hepatocytes, we have shown that the livers with a genetic deletion of Orc2 (or Orc1 and Orc2) undergo endo-reduplication more easily during normal development. Such DNA synthesis is also clearly seen during liver regeneration after partial hepatectomy in the Orc2-/- or Orc1-/-, Orc2-/- livers. Endoreduplication differs from normal replication in that although the cycles of DNA replication (S phase) are separated by a G1 like phase (G phase) they are not separated by an intervening mitosis and cytokinesis, and is commonly seen during normal development in plants and animals (Shu et al., 2018). However, all evidence suggests that the biochemical mechanism of initiating DNA replication is the same in normal mitotic replication and endoreduplication. For example, cyclin E/cdk kinase activity and MCM2-7 protein association with the chromatin oscillate between G phase and S phase) during Drosophila endoreduplication (Lilly and Duronio, 2005) (Su and O’Farrell, 1998). Similarly, in the mouse, cyclin E and CDC6 promote, while geminin inhibits endoreduplication (Welch, 1992) (Bermejo et al., 2002, Gonzalez et al., 2006) (de Renty et al., 2014). Thus, the persistence, and in fact acceleration, of endoreduplication in the mouse livers in the absence of ORC2 and/or ORC1 suggests that there must be an alternate way to recruit sufficient MCM2-7 to the chromatin to support at least the three cycles of endoreduplication required to produce a 16 N nucleus from a 2 N embryonic nucleus during normal development, and to support about two cycles of endoreduplication during liver regeneration.

The acceleration and persistence of endoreduplication even when both Orc1 and Orc2 are genetically inactivated, makes it unlikely that a make-shift ORC-like complex is being assembled for loading the MCM2-7 proteins when two out of the five subunits in the ring are missing. Our results are virtually identical to what was observed when we conditionally deleted Orc1 (Okano-Uchida et al., 2018). There too, endoreduplication in the liver cells not only persisted but was induced prematurely during development. One possible explanation of the greater endoreduplication in both our papers is that mitosis may be arrested earlier in development by G2 DNA damage checkpoints activated by incomplete licensing and replication of the genome in the absence of ORC. As a result, endoreduplication cycles could begin earlier in development resulting in greater endoreduplication.

Is it possible that incomplete deletion of Orc2 or Orc1 genes in the hepatocytes allows enough liver cells to still carry an Orc2 or Orc1 gene to support DNA replication? This is very unlikely, because the in vitro DNA replication experiment with hepatocytes shows that although the ORC2 protein is virtually undetectable in immunoblots and there is evidence of Cre activity in 100% of the hepatocytes, the number of nuclei incorporating EdU in culture are decreased to only 30% the wild type level.

Furthermore, in the partial hepatectomy experiments, nearly 100% of the hepatocytes are positive for EYFP and no ORC2 (or ORC1 protein in the DKO) is detectable in immunoblots of isolated hepatocytes, suggesting high penetrance in the expression of the Cre recombinase in the hepatocytes. Yet the liver succeeds in endoreduplicating to reach nearly normal liver size and EdU incorporation is seen in 35% (Fig. 5H) and 15% (Fig. 7G) of hepatocytes from the Orc2 KO and DKO mice, respectively.

Although the liver size in the Alb-Cre, Orc2f/f mice is not significantly decreased relative to body weight, the liver function tests suggest some impairment of liver function. We cannot yet attribute this deficit of liver function to the decrease in number of cells, or to excess endoreduplication, because ORC subunits are also known to be important for epigenetic control of gene expression (Vermeulen et al., 2010). Future experiments will determine whether the liver pathology could be secondary to epigenetic dysregulation of genes important for liver function. The double knockout female mice are sicker, and here some of the explanation may lie in the smaller liver, but here again epigenetic dysregulation in the absence of ORC cannot be ruled out as a potential cause for the morbidity.

Interestingly, the deletion of Orc2 or of Orc1+Orc2, in the hepatocytes does not consistently increase or decrease any of the other ORC subunits or the examined proteins downstream of ORC like CDC6 or two of the subunits of MCM2-7. Thus, a change in any of these proteins is unlikely to be the explanation for how the hepatocytes license enough origins to support replication during development and regeneration in the absence of the ORC holocomplex. The hepatocyte studies say that the cancer cells are not unique in their ability to bypass the requirement of two ORC subunits. Even primary cells can sometimes load enough MCM2-7 to replicate DNA in the absence of detectable amounts of the ORC holocomplex. It is also worth noting that in cancer cells in culture, we are seeing a near normal level of loading of MCM2-7 on chromatin when Orc1, Orc2 or Orc5 genes are deleted, and that 60% of the origins of replication remain at the same sites as in WT cells (Shibata et al., 2024). Although it is impossible to rule out that a very small amount of ORC protein somehow persists in the hepatocytes (or the cancer cell lines) with these mutations and that this is sufficient to facilitate loading of enough MCM2-7 to support DNA synthesis in vivo or in vitro, this is becoming progressively unlikely.